---
title: "Passerine Body Plan Evolution, Part 3: Shift Timing and Distribution Fits"
colab-packages: "univariateML, evd"
output:
  rmarkdown::html_vignette: default
  rmarkdown::pdf_document:
    latex_engine: xelatex
vignette: >
  %\VignetteIndexEntry{Passerine Body Plan Evolution, Part 3: Shift Timing and Distribution Fits}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include = FALSE}
library(bifrost)
library(classInt)
library(RColorBrewer)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

vignette_figure_caption <- function(html, latex) {
  pandoc_to <- knitr::opts_knit$get("rmarkdown.pandoc.to")
  if (is.character(pandoc_to) && length(pandoc_to) == 1L && grepl("^html", pandoc_to)) html else latex
}
```
```{r, echo=FALSE, results='asis'}
if (knitr::is_html_output()) cat('<style>
/* Remove any default borders/shadows around vignette/pkgdown images */
img, .figure img, .figure > p > img {
  border: 0 !important;
  box-shadow: none !important;
  outline: none !important;
  background: transparent !important;
}

p.caption, .caption {
  border-left: 0.32rem solid #d8dde3;
  color: #2f3439;
  font-size: inherit;
  line-height: inherit;
  margin: 0 3.8% 1.05rem 3.8%;
  padding: 0 0 0 1rem;
  text-align: left !important;
}

p.caption strong, .caption strong {
  font-weight: 700;
}

code {
  overflow-wrap: anywhere;
  word-break: break-word;
}

div.sourceCode,
pre.sourceCode,
pre,
pre code,
code.sourceCode,
pre > code.sourceCode,
pre > code.sourceCode > span {
  max-width: 100%;
  white-space: pre-wrap !important;
  overflow-wrap: anywhere !important;
  word-break: break-word !important;
}

div.sourceCode,
pre {
  overflow-x: auto;
}

pre > code.sourceCode > span {
  display: inline !important;
  text-indent: 0 !important;
  padding-left: 0 !important;
}

@media (max-width: 640px) {
  img, .figure img, .figure > p > img {
    width: 100% !important;
    height: auto !important;
  }
}
</style>
')
```

## Introduction

Accepted shifts become interpretable once they are turned into explicit events: where did the regime change occur, did it increase or decrease the fitted BMM rate, how are fitted regime rates distributed, and how much time separates successive shifts? Part 3 answers those timing and distribution questions from the compact focal passerine body-plan search.

The workflow starts from the same fitted multi-regime Brownian motion model (BMM) used in Parts 1 and 2. The compact focal object contains the mapped tree, fitted regime rates, accepted shifts, and `ic_weights` needed for these summaries; the full Berv et al. (2026) archive remains the source for proposal history, sensitivity searches, and candidate model fits. Part 4 then uses the transition table introduced here to compare shift magnitudes, and Part 5 moves to post-hoc covariance structure.

## Loading the compact focal result

```{r load_compact_bodyplan_search_part3}
# Load the same compact focal result inspected in Part 1.
search_path <- system.file(
  "extdata",
  "avian-skeleton",
  "passerine_bodyplan_search_compact.RDS",
  package = "bifrost"
)
stopifnot(nzchar(search_path))

bodyplan_search <- readRDS(search_path)

# Preserve the S3 class if an older serialized copy lacks it.
if (!inherits(bodyplan_search, "bifrost_search")) {
  class(bodyplan_search) <- c("bifrost_search", class(bodyplan_search))
}

# Confirm that the object is a fitted BMM search with inferred shifts.
data.frame(
  inherits_bifrost_search = inherits(bodyplan_search, "bifrost_search"),
  model = bodyplan_search$model_no_uncertainty$model,
  fitted_regimes = length(bodyplan_search$model_no_uncertainty$param),
  accepted_shifts = length(bodyplan_search$shift_nodes_no_uncertainty)
)
```

For `bifrost_search` inputs, the functions in this vignette check that the fitted model is a multi-regime BMM before computing rate-based summaries. That guard matters because the workflow depends on a mapped heterogeneous-rate history, not just any fitted comparative model object.

## Extracting shift transitions

Berv et al. (2026) also summarized the inferred regime shifts themselves. `shift_transitions()` extracts parent-child state changes from the final mapped tree and annotates each transition with the parent and child regime rates.

```{r focal_shift_transitions_part3}
# Extract accepted shifts, excluding the root state assignment.
bodyplan_transitions <- shift_transitions(bodyplan_search, include_root = FALSE)

# Preview timing, state changes, and direction of rate change.
head(bodyplan_transitions[, c(
  "node",
  "age",
  "parent_state",
  "child_state",
  "rate_change"
)])
```

The `rate_change` column bridges the fitted BMM object and the shift-magnitude summaries in Part 4. Here we keep the transition table to orient shift timing and state changes; Part 4 counts directions and compares `rate_delta` magnitudes.

## Annotating shift nodes on the rate tree

The transition table also lets you annotate the branch-rate arc tree from Part 1. Berv et al. (2026) used this layer to flag rate-change nodes on the branch-rate map. The branch-rate view uses `rateMap()` and `rateMapView()` directly; the public annotation workflow here is `shift_node_marks()` followed by its S3 `plot()` method. The first step prepares the node marks, letter key, and low-support flags without touching the active plot.

```{r focal_shift_annotation_tree_view_part3, message=FALSE}
# Build the branch-rate map used beneath the shift-node annotations.
bodyplan_rate_map <- rateMap(bodyplan_search, progress = FALSE)

# Recreate the manuscript's Fisher-binned log-rate scale and color palette.
bodyplan_rate_breaks <- classInt::classIntervals(
  log(bodyplan_search$model_no_uncertainty$param),
  n = 10,
  style = "fisher"
)$brks

bodyplan_rate_breaks[c(1, length(bodyplan_rate_breaks))] <-
  range(bodyplan_rate_map$intervals$value)

bodyplan_rate_palette <- grDevices::colorRampPalette(
  rev(RColorBrewer::brewer.pal(11, "RdYlBu"))
)(length(bodyplan_rate_breaks) - 1L)

# Apply those fixed categories to a reusable tree view.
bodyplan_shift_tree <- rateMapView(
  bodyplan_rate_map,
  color_mode = "category",
  category_breaks = bodyplan_rate_breaks,
  palette = bodyplan_rate_palette,
  legend_title = "Log fitted BMM rate"
)
```

```{r focal_shift_annotation_data_part3}
# Prepare shift-node marks without plotting them yet.
bodyplan_shift_marks <- shift_node_marks(
  bodyplan_transitions,
  ic_weights = bodyplan_search$ic_weights,
  marker_base_cex = 1.5,
  marker_scale_factor = 0.08,
  marker_transform = "square_root",
  letter_order = "child_state"
)

bodyplan_shift_marks$summary
```

The returned `shift_node_marks` object is reusable: inspect its tables first, then pass it to `plot()` to draw on an already plotted tree. The letter key below follows the same increase-node labeling used by Berv et al. (2026): increase nodes are ordered by child regime state, then labeled A-R. The `percentage_change` column is the signed percent change from the parent regime rate to the child regime rate.

```{r focal_shift_annotation_key_part3}
# Keep the key compact: only lettered increase nodes need decoding.
bodyplan_shift_marks$increase_key
```

```{r focal_shift_annotation_tree_part3, message=FALSE, fig.align="center", fig.width=8.4, fig.height=5.2, dpi=220, fig.retina=1, out.width="100%", fig.cap=vignette_figure_caption("**Figure 1. Focal rate-increase nodes overlaid on the Berv et al. branch-rate tree.** Branch colors use the same Fisher-binned reversed RdYlBu log-rate palette as Berv et al. (2026). White lettered circles mark inferred shifts to faster fitted BMM regimes, with marker size following the square-root scaling of percent rate change used in that figure.", "Focal rate-increase nodes overlaid on the Berv et al. branch-rate tree. Branch colors use the same Fisher-binned reversed RdYlBu log-rate palette as Berv et al. (2026). White lettered circles mark inferred shifts to faster fitted BMM regimes, with marker size following the square-root scaling of percent rate change used in that figure.")}
# Draw the rate tree first so the shift markers can be overlaid in tree coordinates.
plot(
  bodyplan_shift_tree,
  type = "arc",
  show_tip_labels = FALSE,
  lwd = c(1.20, 5.0),
  legend = 42,
  legend_fsize = 0.72,
  legend_digits = 3,
  arc_height = 0.75,
  mar = c(0.2, 0.2, 0.2, 0.2)
)

# Add only supported rate-increase markers to the active tree plot.
plot(
  bodyplan_shift_marks,
  rate_changes = "increase",
  show_low_support = FALSE,
  show_legend = FALSE
)
```

## Fitting rate distributions

`fit_rate_distribution()` compares the fitted BMM regime rates across candidate continuous models. It accepts the `bifrost_search` object directly, log-transforms the positive regime rates by default, and ranks a broad default set of `univariateML` candidates.

```{r focal_rate_distribution_part3, message=FALSE}
# Fit candidate distributions to log fitted BMM regime rates.
bodyplan_rate_fit <- fit_rate_distribution(bodyplan_search)

# Precompute Gumbel bootstrap draws once; this also stores the D_KL summary.
bodyplan_rate_fit <- bootstrap_rate_distribution(
  bodyplan_rate_fit,
  model = "gumbel",
  reps = 1000,
  seed = 2026
)

# Pull out rankings, selected parameters, and the Gumbel goodness-of-fit summary.
rate_rankings <- as.data.frame(bodyplan_rate_fit, component = "rankings")
rate_parameters <- as.data.frame(
  bodyplan_rate_fit,
  component = "parameters",
  models = bodyplan_rate_fit$rankings$ml[seq_len(min(3, nrow(bodyplan_rate_fit$rankings)))]
)
rate_goodness <- as.data.frame(bodyplan_rate_fit, component = "goodness")

# Print compact previews; full tables remain in the objects above.
rate_rankings_preview <- rate_rankings[seq_len(min(5, nrow(rate_rankings))), ]
rate_parameters_preview <- rate_parameters[
  rate_parameters$selected,
  !names(rate_parameters) %in% c("selected_AIC", "selected_BIC"),
  drop = FALSE
]

print(rate_rankings_preview)
cat("\n\n")
print(rate_parameters_preview)
cat("\n\n")
print(rate_goodness)
```

The preview keeps the top candidate distributions, parameter estimates for the selected Gumbel fit, and the bootstrapped Kullback-Leibler divergence (`D_KL`) for the Gumbel overlay. Use `as.data.frame(bodyplan_rate_fit, component = "rankings")` or `component = "parameters"` to inspect the complete model-selection tables.

The histogram below reuses the Fisher breaks and branch-rate palette from Figure 1. `fit_rate_distribution()` stores the fitted candidates and rankings; `bootstrap_rate_distribution()` adds the reproducible Gumbel bootstrap draws that `plot()` displays.

```{r focal_gumbel_distribution_plot_part3, message=FALSE, fig.align="center", fig.width=7.2, fig.height=4.6, dpi=220, fig.retina=1, out.width="100%", fig.cap=vignette_figure_caption("**Figure 2. Fitted Gumbel distribution for the focal BMM regime rates.** Histogram bars use Fisher breaks and the reversed RdYlBu palette used for branch-rate bins in Berv et al. (2026). The red line is the maximum-likelihood Gumbel density, the gray dashed curve is the empirical density, tick marks show fitted regime rates, the pale band is the pointwise 95% bootstrap interval, and thin red curves show bootstrap density draws.", "Fitted Gumbel distribution for the focal BMM regime rates. Histogram bars use Fisher breaks and the reversed RdYlBu palette used for branch-rate bins in Berv et al. (2026). The red line is the maximum-likelihood Gumbel density, the gray dashed curve is the empirical density, tick marks show fitted regime rates, the pale band is the pointwise 95\\% bootstrap interval, and thin red curves show bootstrap density draws.")}
# Plot the fitted Gumbel density with its stored bootstrap uncertainty.
plot(
  bodyplan_rate_fit,
  model = "gumbel",
  bootstrap = TRUE,
  bootstrap_curves = 120,
  breaks = bodyplan_rate_breaks,
  hist_col = bodyplan_rate_palette
)
```

## Waiting times between shifts

`shift_waiting_times()` computes the waiting-time views used to separate tree-wide burstiness from lineage-level shift histories. Berv et al. (2026) report two distributional summaries: whole-tree chronological waiting times between successive shifts, and root-to-tip lineage waiting times used for per-lineage candidate fits. The acceleration summary isolates increase-to-increase gaps; we print it as context, not as the primary reported waiting-time distribution.

```{r focal_shift_waiting_times_part3}
# Compute global, lineage-level, and acceleration waiting-time summaries.
bodyplan_waiting_times <- shift_waiting_times(bodyplan_search)

# Print the compact summaries rather than the full lineage table.
bodyplan_waiting_times$summaries$global
bodyplan_waiting_times$summaries$lineage
bodyplan_waiting_times$acceleration$summary
```

In Berv et al. (2026), the whole-tree chronological waits support the Weibull result for shift timing across the phylogeny, while the lineage waits support the per-lineage Exponential result for tip-to-root paths.

## Fitting waiting-time distributions

The first fit below repeats the whole-phylogeny comparison from Berv et al. (2026): chronological waiting times between successive shifts are compared across the same positive-support candidate models.

```{r focal_waiting_time_distribution_part3, message=FALSE}
# Fit positive-support candidates to whole-tree chronological waiting times.
# The default includes the initial root-to-first-shift interval.
bodyplan_waiting_fit <- fit_waiting_time_distribution(
  bodyplan_waiting_times,
  global_include_root_wait = TRUE
)

# Extract rankings plus parameters for the top-supported candidates.
waiting_rankings <- as.data.frame(bodyplan_waiting_fit, component = "rankings")
waiting_parameters <- as.data.frame(
  bodyplan_waiting_fit,
  component = "parameters",
  models = bodyplan_waiting_fit$rankings$ml[seq_len(min(3, nrow(bodyplan_waiting_fit$rankings)))]
)

# Print compact previews; full tables remain in the fitted object.
waiting_rankings_preview <- waiting_rankings[
  seq_len(min(5, nrow(waiting_rankings))),
  ,
  drop = FALSE
]
waiting_parameters_preview <- waiting_parameters[
  waiting_parameters$selected,
  !names(waiting_parameters) %in% c("selected_AIC", "selected_BIC"),
  drop = FALSE
]

print(waiting_rankings_preview)
cat("\n\n")
print(waiting_parameters_preview)
```

Set `global_include_root_wait = FALSE` when you need to mirror the archived `TimeToNext[-1]` check rather than the reported whole-tree Weibull summary.

Setting `by_lineage = TRUE` repeats the candidate-model comparison for each lineage with enough waiting-time observations. These fits provide the compact-data check for the Berv et al. (2026) statement that most tip-to-root paths favor a constant-rate Exponential waiting-time model.

```{r focal_lineage_waiting_distribution_part3, message=FALSE}
# Repeat the waiting-time fits independently for each lineage.
bodyplan_lineage_waiting_fit <- fit_waiting_time_distribution(
  bodyplan_waiting_times,
  by_lineage = TRUE
)

# Summarize how many lineages had enough waiting times to fit.
lineage_summary <- as.data.frame(bodyplan_lineage_waiting_fit, component = "lineage")

lineage_fit_counts <- data.frame(
  lineages_with_waits = nrow(lineage_summary),
  fitted_lineages = sum(lineage_summary$fit_success),
  skipped_lineages = sum(!lineage_summary$fit_success)
)

fitted_lineage_summary <- lineage_summary[lineage_summary$fit_success, ]

lineage_exponential_check <- data.frame(
  criterion = c("BIC", "AIC"),
  best_model = "exp",
  exponential_lineages = c(
    sum(fitted_lineage_summary$best_BIC == "exp"),
    sum(fitted_lineage_summary$best_AIC == "exp")
  ),
  fitted_lineages = nrow(fitted_lineage_summary),
  percent_fitted = c(
    round(
      100 * sum(fitted_lineage_summary$best_BIC == "exp") /
        nrow(fitted_lineage_summary),
      1
    ),
    round(
      100 * sum(fitted_lineage_summary$best_AIC == "exp") /
        nrow(fitted_lineage_summary),
      1
    )
  ),
  reported_percent = c("86%", NA)
)

# Print the compact summaries needed for the manuscript check.
print(lineage_fit_counts)
cat("\n\n")
print(lineage_exponential_check)
cat("\n\n")
print(bodyplan_lineage_waiting_fit$lambda_summary)
```

The reported statement that 86% of tip-to-root paths favor a constant-rate Exponential model comes from the BIC-supported per-lineage model summary in the archived workflow. The compact object gives 1,482 BIC Exponential winners among 1,708 fitted lineage paths, or 86.8%, while AIC gives a higher count of 1,603 Exponential winners, or 93.9%. We print both criteria to make that distinction explicit; the small difference between the compact BIC percentage and the published 86% reflects reported-value rounding/truncation or full-archive run details.

The full per-lineage model-count and parameter tables remain available with `as.data.frame(bodyplan_lineage_waiting_fit, component = "lineage")` and `component = "lineage_parameters"`; here we print only the compact summaries needed to check the reported 86% statement. The `lambda_summary` table includes the mean, median, standard deviation, and a 95% shortest-interval summary for the lineage-level Exponential rate parameter.

## Practical Takeaways

- `shift_transitions()` extracts fitted parent-child regime changes from the focal BMM map and records timing, direction, and signed rate differences.
- `rateMap()` provides the Berv et al. (2026) branch-rate tree; `shift_node_marks()` can optionally use drop-one `ic_weights` to flag low-support shifts.
- `shift_node_marks()` prepares rate-increase annotations, and `plot()` overlays them on an existing tree plot without rebuilding the branch-rate layer.
- `fit_rate_distribution()` wraps `univariateML::model_select()` for fitted BMM regime rates and `plot(rate_fit, model = "gumbel")` draws the Berv et al. (2026)-style Gumbel overlay.
- `shift_waiting_times()` computes global, lineage-level, and acceleration-focused waiting-time summaries from the compact focal object.
- `fit_waiting_time_distribution()` supports both whole-tree chronological waits and per-lineage waiting-time fits used to check the Berv et al. (2026) Weibull and Exponential summaries.
- Part 4 uses `compare_shift_magnitudes()` and the compact sensitivity bundle to compare log absolute rate-change magnitudes for increases and decreases.
- Part 5 reuses the temporal regime framework for post-hoc integration summaries, not for spatial/local assemblage covariance.
- Use the complete Berv et al. (2026) archive for proposal-level history, candidate model fits, and external ecological metadata; use the compact focal object for these timing and distribution summaries.

## References

If you use these data or reproduce this workflow, the most relevant citations are:

- Berv, Jacob S., Charlotte M. Probst, Santiago Claramunt, J. Ryan Shipley, Matt Friedman, Stephen A. Smith, David F. Fouhey, and Brian C. Weeks. 2026. "Rates of passerine body plan evolution in time and space." *Nature Ecology & Evolution*. <https://doi.org/10.1038/s41559-026-03110-5>
- Berv, Jacob S., Charlotte M. Probst, Santiago Claramunt, J. Ryan Shipley, Matt Friedman, Stephen A. Smith, David F. Fouhey, and Brian C. Weeks. 2026. "Supplementary data archive for Rates of passerine body plan evolution in time and space" (v1.0.0) [Data set]. Zenodo. <https://doi.org/10.5281/zenodo.19198393>
- Claramunt, Santiago, Christopher Sheard, Joseph W. Brown, Guillermo Cortes-Ramirez, Joel Cracraft, Mason M. Su, Brian C. Weeks, and Joseph A. Tobias. 2025. "A new time tree of birds reveals the interplay between dispersal, geographic range size, and diversification." *Current Biology*. Advance online publication. <https://doi.org/10.1016/j.cub.2025.07.004>

## Software Used in This Vignette

- `bifrost` for transition extraction, shift waiting times, rate-map helpers, and distribution-fitting wrappers.
- `ape` and `phytools` underlie the mapped-tree calculations in these summaries.
- `classInt` computes the Fisher breaks for the rate-distribution histogram.
- `RColorBrewer` supplies the Berv et al. (2026) branch-rate palette.
- `univariateML` fits candidate distributions when these examples run.
- `evd` evaluates the fitted Gumbel density for the rate-distribution overlay.

## AI Assistance

This vignette was developed with assistance from OpenAI tools for drafting, editing, and figure refinement; all scientific content, interpretation, and final decisions were reviewed by the authors.
