---
title: "bifrost background: model, search strategy, and interpretation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{bifrost background: model, search strategy, and interpretation}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

`bifrost` is designed for branch-level inference of multivariate evolutionary rate variation across phylogenies. Its core use-case is identifying hidden regime shifts in trait evolution while fitting multivariate models directly in trait space.

In practical terms, `bifrost` asks:

- **Where** on the tree are shifts supported?
- **When** do those shifts occur?
- **How** does covariance/rate structure differ across inferred regimes?

The current implementation searches for shifts under a multi-rate Brownian Motion (BMM) model with proportional regime variance-covariance (VCV) scaling.

## Why this framework?

Many comparative datasets are now high-dimensional (e.g., geometric morphometric coordinates), where `p` (number of trait variables) can approach or exceed `n` (number of taxa). `bifrost` uses penalized-likelihood multivariate GLS fitting (`mvMORPH::mvgls`) to keep model fitting tractable in these settings.

Key implications:

- traits are modeled jointly (rather than reduced to a few PCs),
- rate/covariance heterogeneity can be inferred on branches,
- fitted models are conditioned on inferred regime structure.

## Data requirements

At minimum, provide:

1. A phylogeny (branch lengths interpreted in units of time; ultrametricity is not required).
2. A multivariate trait matrix with taxa as row names.

Important checks before fitting:

- `rownames(trait_data)` must match `tree$tip.label` exactly,
- trait ordering must align with tree tip ordering,
- branch lengths should be biologically meaningful for comparative interpretation.

## Search strategy in `searchOptimalConfiguration()`

`searchOptimalConfiguration()` performs a greedy, step-wise search:

1. Fit a baseline (single-regime) model.
2. Generate one-shift candidate trees given `min_descendant_tips`.
3. Score candidates (optionally in parallel).
4. Iteratively accept shifts that improve IC by at least `shift_acceptance_threshold`.
5. Optionally compute per-shift IC support by drop-one refits.

The accepted shift set defines the optimal no-uncertainty configuration under the selected criterion (`"GIC"` or `"BIC"`).

### High-level workflow

```mermaid
flowchart TD

subgraph S[Setup]
  A([Start]) --> B[Validate tree and traits]
  B --> C[Paint baseline regime 0]
  C --> D[Generate one-shift candidates]
end

subgraph F[Fitting and scoring]
  D --> E[Fit baseline mvgls]
  E --> G[Compute baseline IC]
  G --> H[Score candidates]
end

subgraph G2[Greedy updates]
  H --> I{Candidate improves IC by threshold?}
  I -- Yes --> J[Accept shift]
  I -- No --> K[Reject shift]
  J --> L{More candidates?}
  K --> L
  L -- Yes --> H
  L -- No --> M[Finalize optimal model]
end

subgraph U[Uncertainty (optional)]
  M --> N{Compute per-shift IC weights?}
  N -- Yes --> O[Drop-one shift refits]
  N -- No --> P[Return result]
  O --> P
end
```

## Understanding key arguments

Commonly tuned parameters include:

- `min_descendant_tips`: minimum clade size for candidate shifts.
- `shift_acceptance_threshold`: required IC improvement for acceptance.
- `IC`: model comparison metric (`"GIC"` or `"BIC"`).
- `num_cores`: parallel candidate scoring.
- `formula`, `method`, `penalty` (via `mvgls`): model-fitting assumptions.

For exploratory analyses, start conservative (higher thresholds, larger clade-size minimum), then perform sensitivity analyses across parameter settings.

## Outputs and interpretation

The return object from `searchOptimalConfiguration()` includes:

- optimal SIMMAP tree with accepted shifts,
- final fitted `mvgls` model,
- inferred shift node IDs,
- baseline and optimal IC values,
- regime-specific VCV estimates,
- optional per-shift IC weights/evidence ratios,
- optional model fit history.

Interpretation guidance:

- large IC gains from baseline indicate strong support for heterogeneity,
- supported shifts indicate branch-localized changes in evolutionary rate/covariance,
- uncertainty weights help prioritize which accepted shifts are most influential.

## Performance and scalability

For large trees or high-dimensional traits:

- use `future`/`future.apply` parallelization,
- disable plotting during fitting (`plot = FALSE`),
- increase `min_descendant_tips` to reduce candidate space,
- use stricter acceptance thresholds to reduce false positives,
- replicate analyses across seeds/thresholds for robustness.

Example parallel plan:

```{r}
library(future)
plan(multisession)  # or multicore on Linux/macOS
```

## Positioning relative to multivariate pGLS

Although developed for regime-shift inference, `bifrost` can also be used in multivariate pGLS-style analyses with predictors. In that setting, the fitted regression is conditioned on an inferred residual covariance structure that allows hidden branch-specific rate variation.

Conceptually, this is related to hidden-state comparative approaches, but with regimes acting on variance/rate structure rather than introducing regime-specific means.

## Reproducibility recommendations

For transparent analyses:

- report `sessionInfo()`,
- record package versions (especially `mvMORPH`),
- consider project dependency locking (e.g., `renv`),
- archive trees, trait matrices, and script parameters used for final fits.

## Where to go next

- Applied tutorial with empirical data:
  `vignette("jaw-shape-vignette", package = "bifrost")`
- Function documentation:
  `?searchOptimalConfiguration`
- Citation information:
  `citation("bifrost")`
