---
title: "Assessing Expected Performance with Dataset-Matched Simulations"
output:
  rmarkdown::html_vignette: default
  rmarkdown::pdf_document:
    latex_engine: xelatex
vignette: >
  %\VignetteIndexEntry{Assessing Expected Performance with Dataset-Matched Simulations}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

pkg_file <- function(...) {
  installed_path <- system.file(..., package = "bifrost")
  if (nzchar(installed_path)) {
    return(installed_path)
  }

  candidates <- c(
    file.path("inst", ...),
    file.path("..", "inst", ...)
  )

  for (path in candidates) {
    if (file.exists(path)) {
      return(path)
    }
  }

  stop("Could not locate: ", file.path(...))
}
```

## Introduction

This vignette ports the manuscript simulation workflow into a package-generalized form. The idea is simple: before fitting an empirical `bifrost` search, we simulate datasets that resemble the tree shape and trait structure of the real dataset, then ask how the search behaves under three scenarios:

1. a null model with no shifts, to estimate expected false positives,
2. the current generating model, where shifts are proportional covariance scalings, and
3. a deliberately non-generating robustness scenario, where trait correlations evolve in ways that violate strict proportional scaling.

That workflow mirrors the simulation framework used in the passerine body-plan manuscript, but the new package API is designed so that a new user can run the same style of assessment on their own dataset.

## Setup

```{r load_packages, eval = FALSE}
library(bifrost)
library(ape)
```

## Loading a dataset and building a simulation template

The packaged passerine skeleton example is a convenient demonstration because it shows the formula-based workflow: 12 skeletal traits are modeled jointly while log body mass is kept as a phylogenetic covariate.

```{r load_bodyplan_example}
tree_path <- pkg_file("extdata", "avian-skeleton", "passerine_bodyplan_tree.tre")
trait_path <- pkg_file("extdata", "avian-skeleton", "passerine_bodyplan_data.RDS")

bird_tree <- ape::read.tree(tree_path)
bodyplan_data <- readRDS(trait_path)
bodyplan_data <- as.matrix(bodyplan_data[bird_tree$tip.label, , drop = FALSE])

skeletal_cols <- setdiff(colnames(bodyplan_data), "vertnet_mass")
bodyplan_data <- bodyplan_data[, c(skeletal_cols, "vertnet_mass"), drop = FALSE]

formula_str <- sprintf(
  "trait_data[, 1:%d] ~ trait_data[, %d]",
  length(skeletal_cols),
  ncol(bodyplan_data)
)

c(
  species = nrow(bodyplan_data),
  skeletal_traits = length(skeletal_cols),
  total_columns = ncol(bodyplan_data)
)
```

The next step is to fit a global empirical template. This reproduces the manuscript logic in which simulation parameters were estimated from a single-regime empirical fit before generating null and shifted replicates.

```{r create_template_code, eval = FALSE}
bodyplan_template <- createSimulationTemplate(
  baseline_tree = bird_tree,
  trait_data = bodyplan_data,
  formula = formula_str,
  response_columns = seq_along(skeletal_cols),
  predictor_columns = ncol(bodyplan_data),
  method = "LL",
  error = TRUE
)
```

```{r create_template_print, eval = FALSE}
bodyplan_template
```

The template stores the aligned analysis data, the fitted global `mvgls` model, and the empirical variance/covariance summaries used to generate a fresh ancestral covariance matrix for each simulation replicate.

## A note on manuscript alignment

One manuscript-specific nuance is worth making explicit. In the passerine body-plan analyses, the global empirical model used body mass as a covariate when estimating the overall covariance scale, but the simulation searches themselves were run as intercept-only searches. In other words, the manuscript used the mass-corrected empirical fit to calibrate the scale of the simulations, then asked a narrower question: how conservative and how sensitive is the shift-search itself on a dataset of this size and covariance structure?

That shortcut is best understood as an approximation to the residual multivariate variation the search was operating on, not as a full generative reconstruction of the empirical covariate-plus-shift model. Body mass was used to estimate the residual variance and covariance regime, but the simulation benchmark did not try to simulate predictors or to re-evaluate predictor estimation itself.

That shortcut was reasonable for the manuscript because the simulation study was aimed at calibrating branch-shift detection, not at jointly validating both predictor estimation and shift recovery. For a new user, the safer default is the fully matched workflow shown above, where the simulated datasets preserve the predictor columns and the simulation study uses the same formula you plan to fit empirically.

If you want a closer analogue of the manuscript shortcut, the package can also be used in an intercept-only, response-only mode:

```{r manuscript_style_template, eval = FALSE}
bodyplan_response_only <- bodyplan_data[, skeletal_cols, drop = FALSE]

bodyplan_template_intercept <- createSimulationTemplate(
  baseline_tree = bird_tree,
  trait_data = bodyplan_response_only,
  formula = "trait_data ~ 1",
  method = "LL",
  error = TRUE
)
```

Using `bodyplan_template_intercept` makes the simulation study target the multivariate response block alone. That is closer to the original manuscript search simulations, but it is less faithful to a truly formula-based empirical analysis than the full `bodyplan_template` workflow above.

The same response-only shortcut also works with subset-response intercept formulas such as `"trait_data[, 1:12] ~ 1"`, as long as `response_columns` is supplied so the template knows which columns to simulate.

## Practical guardrails

The current simulation wrappers include a few package-level guardrails that are worth knowing about before launching a large study.

For formula-based workflows, predictor columns are preserved automatically and should stay in the simulated dataset. In other words, once you build a template like `bodyplan_template`, the downstream simulation studies should keep using that same formula-based structure rather than dropping the predictor block.

Reproducibility for replicated studies is controlled by the wrapper-level `seed` argument in `runFalsePositiveSimulationStudy()` and `runShiftRecoverySimulationStudy()`. There is no need to pass a separate seed inside `simulation_options`; the wrappers manage replicate-level seeding internally.

The study wrappers also inherit key fit settings such as `method` and `error` from the simulation template. In many cases you can leave those out of `search_options` unless you intentionally want to override the template defaults.

Finally, it is a good idea to inspect `per_replicate$n_candidates` after a run. If `min_descendant_tips` is set so high that very few candidate shifts remain, the study will be much less informative about expected model performance.

In particular, a replicate with `n_candidates == 0` should be treated as unevaluable rather than as evidence for good recovery performance. In that case, recovery specificity, false-positive rate, and balanced accuracy are reported as `NA`, and the usual fix is to lower `min_descendant_tips` until a meaningful candidate set is available.

## 1. Estimating false positives under the null model

The first manuscript-style question is: if the real dataset had no shifts at all, how often would the chosen search settings infer one anyway?

The code below keeps the full formula-based workflow intact. If you want the closer manuscript analogue instead, replace `bodyplan_template` with `bodyplan_template_intercept` and change the search formula to `"trait_data ~ 1"`.

```{r false_positive_study, eval = FALSE}
fp_study <- runFalsePositiveSimulationStudy(
  template = bodyplan_template,
  n_replicates = 100,
  tree_tip_count = 250,
  simulation_options = list(),
  search_options = list(
    formula = formula_str,
    min_descendant_tips = 10,
    shift_acceptance_threshold = 10,
    IC = "GIC",
    num_cores = 1,
    method = "LL",
    error = TRUE,
    plot = FALSE,
    store_model_fit_history = FALSE
  ),
  num_cores = 8,
  seed = 5
)
```

```{r false_positive_inspect, eval = FALSE}
fp_study
head(fp_study$per_replicate)
fp_study$study_summary
```

In practice, this is the stage where you tune conservative choices such as `shift_acceptance_threshold`, `IC`, and `min_descendant_tips`.

## 2. Assessing recovery under the generating model

The second step mirrors the manuscript's proportional-scaling simulations. Here the generating process matches the current `bifrost` BMM search assumptions: shifted regimes are scalar multiples of the background covariance matrix.

```{r proportional_recovery_study, eval = FALSE}
recovery_prop <- runShiftRecoverySimulationStudy(
  template = bodyplan_template,
  n_replicates = 100,
  tree_tip_count = 250,
  simulation_options = list(
    num_shifts = 5,
    min_shift_tips = 10,
    max_shift_tips = 20,
    scale_mode = "proportional",
    scale_factor_range = c(0.1, 2.0),
    exclude_range = c(0.5, 1.5),
    buffer = 3
  ),
  search_options = list(
    formula = formula_str,
    min_descendant_tips = 10,
    shift_acceptance_threshold = 10,
    IC = "GIC",
    num_cores = 1,
    method = "LL",
    error = TRUE,
    uncertaintyweights_par = TRUE,
    plot = FALSE,
    store_model_fit_history = FALSE
  ),
  fuzzy_distance = 2,
  weighted = TRUE,
  num_cores = 8,
  seed = 5
)
```

```{r proportional_recovery_inspect, eval = FALSE}
recovery_prop
recovery_prop$study_summary
recovery_prop$evaluation$strict
recovery_prop$evaluation$fuzzy
recovery_prop$evaluation$weighted
```

## 3. Assessing robustness under the non-generating scenario

The manuscript also tested a deliberately mis-specified scenario in which trait correlations evolve. This is useful because it asks how well the search behaves when the real data deviate from strict proportional VCV scaling.

```{r correlation_recovery_study, eval = FALSE}
recovery_corr <- runShiftRecoverySimulationStudy(
  template = bodyplan_template,
  n_replicates = 100,
  tree_tip_count = 250,
  simulation_options = list(
    num_shifts = 5,
    min_shift_tips = 10,
    max_shift_tips = 20,
    scale_mode = "correlation",
    scale_factor_range = c(0.1, 2.0),
    exclude_range = c(0.5, 1.5),
    buffer = 3
  ),
  search_options = list(
    formula = formula_str,
    min_descendant_tips = 10,
    shift_acceptance_threshold = 10,
    IC = "GIC",
    num_cores = 1,
    method = "LL",
    error = TRUE,
    uncertaintyweights_par = TRUE,
    plot = FALSE,
    store_model_fit_history = FALSE
  ),
  fuzzy_distance = 2,
  weighted = TRUE,
  num_cores = 8,
  seed = 5
)
```

```{r correlation_recovery_inspect, eval = FALSE}
recovery_corr
recovery_corr$study_summary
recovery_corr$evaluation$fuzzy
```

This scenario is not the current generating model for `bifrost`, so the goal is not perfect recovery. Instead, it is a robustness check: do the same search settings still behave reasonably when the data depart from the assumed model in a biologically plausible way?

## 4. Tuning GIC and BIC separately

In practice, it is often cleaner to treat GIC and BIC as separate workflows rather than as one more parameter in a single giant grid. The new tuning helpers are designed around that idea: fix `IC = "GIC"` and tune the threshold/clade-size settings within GIC, then repeat the same process for BIC.

The helper below runs the same three-study logic shown above for every combination of `shift_acceptance_threshold` and `min_descendant_tips`, but it holds `IC` fixed throughout the grid.

```{r tune_gic_grid, eval = FALSE}
gic_grid <- runSearchTuningGrid(
  template = bodyplan_template,
  IC = "GIC",
  shift_acceptance_thresholds = c(2, 10, 20),
  min_descendant_tips_values = c(10, 15),
  tree_tip_count = 250,
  null_replicates = 100,
  recovery_replicates = 100,
  proportional_simulation_options = list(
    num_shifts = 5,
    min_shift_tips = 10,
    max_shift_tips = 20,
    scale_factor_range = c(0.1, 2.0),
    exclude_range = c(0.5, 1.5),
    buffer = 3
  ),
  base_search_options = list(
    formula = formula_str,
    method = "LL",
    error = TRUE,
    num_cores = 1,
    uncertaintyweights_par = TRUE,
    plot = FALSE,
    store_model_fit_history = FALSE
  ),
  weighted = TRUE,
  num_cores = 8,
  seed = 5
)
```

```{r select_gic_grid, eval = FALSE}
gic_tuned <- selectTunedSearchParameters(
  gic_grid,
  max_false_positive_rate = 0.05,
  max_any_false_positive = 0.20,
  min_evaluable_fraction = 0.8,
  primary_metric = "weighted_fuzzy_f1",
  tie_break = "conservative"
)

gic_tuned$selected_row
gic_tuned$recommended_search_options
```

The same pattern applies to BIC:

```{r tune_bic_grid, eval = FALSE}
bic_grid <- runSearchTuningGrid(
  template = bodyplan_template,
  IC = "BIC",
  shift_acceptance_thresholds = c(2, 10, 20),
  min_descendant_tips_values = c(10, 15),
  tree_tip_count = 250,
  null_replicates = 100,
  recovery_replicates = 100,
  proportional_simulation_options = list(
    num_shifts = 5,
    min_shift_tips = 10,
    max_shift_tips = 20,
    scale_factor_range = c(0.1, 2.0),
    exclude_range = c(0.5, 1.5),
    buffer = 3
  ),
  base_search_options = list(
    formula = formula_str,
    method = "LL",
    error = TRUE,
    num_cores = 1,
    uncertaintyweights_par = TRUE,
    plot = FALSE,
    store_model_fit_history = FALSE
  ),
  weighted = TRUE,
  num_cores = 8,
  seed = 5
)

bic_tuned <- selectTunedSearchParameters(
  bic_grid,
  max_false_positive_rate = 0.05,
  max_any_false_positive = 0.20,
  min_evaluable_fraction = 0.8,
  primary_metric = "weighted_fuzzy_f1",
  tie_break = "conservative"
)
```

At that point you have two tuned workflows:

1. the best threshold/clade-size combination within GIC, and
2. the best threshold/clade-size combination within BIC.

That is usually a more interpretable outcome than one global optimization across both IC families.

## 5. Carry the tuned settings into the empirical search

The selected search options are ready to pass directly into `searchOptimalConfiguration()`. For example, if you decide to proceed with the GIC-tuned workflow:

```{r empirical_search_after_simulations, eval = FALSE}
empirical_result <- do.call(
  searchOptimalConfiguration,
  utils::modifyList(
    gic_tuned$recommended_search_options,
    list(
      baseline_tree = bird_tree,
      trait_data = bodyplan_data,
      num_cores = 20,
      store_model_fit_history = TRUE,
      verbose = TRUE
    )
  )
)
```

## Intercept-only datasets

The same workflow also applies to intercept-only datasets such as GPA-aligned landmark matrices. In that case the template call is simpler because the multivariate response is just the full trait matrix:

```{r intercept_only_template, eval = FALSE}
jaw_template <- createSimulationTemplate(
  baseline_tree = fish_tree,
  trait_data = fish_data,
  formula = "trait_data ~ 1",
  method = "H&L",
  error = TRUE
)
```

The downstream null, proportional, and correlation-scenario studies are then identical in structure; only the template and search options change.

## Takeaway

The new simulation API makes the manuscript assessment workflow reusable:

1. fit a global empirical template,
2. estimate false positives under the null model,
3. measure recovery under the generating model,
4. test robustness under the non-generating correlation scenario, and
5. tune GIC and BIC as separate fixed-IC workflows, then
6. choose empirical search settings with those results in hand.

That is the recommended way to calibrate `bifrost` for a new dataset whose tree size, trait dimension, and predictor structure may differ substantially from the original manuscript example.
