---
title: "Calibrating bifrost with Dataset-Matched Simulations"
output:
  rmarkdown::html_vignette: default
  rmarkdown::pdf_document:
    latex_engine: xelatex
vignette: >
  %\VignetteIndexEntry{Calibrating bifrost with Dataset-Matched Simulations}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  out.width = "100%",
  dpi = 170,
  fig.retina = 2,
  dev = "png"
)

Sys.setenv(
  OMP_NUM_THREADS = "1",
  OPENBLAS_NUM_THREADS = "1",
  MKL_NUM_THREADS = "1",
  VECLIB_MAXIMUM_THREADS = "1",
  BLIS_NUM_THREADS = "1",
  NUMEXPR_NUM_THREADS = "1"
)

if (!exists("createSimulationTemplate", mode = "function")) {
  pkg_root <- if (file.exists("DESCRIPTION")) {
    "."
  } else if (file.exists(file.path("..", "DESCRIPTION"))) {
    ".."
  } else {
    NULL
  }

  if (!is.null(pkg_root) && requireNamespace("pkgload", quietly = TRUE)) {
    pkgload::load_all(pkg_root, quiet = TRUE, export_all = FALSE, helpers = FALSE)
  } else {
    library(bifrost)
  }
}

render_preview_table <- function(x, caption, col_names = colnames(x), digits = 3) {
  x <- as.data.frame(x, stringsAsFactors = FALSE, check.names = FALSE)

  if (length(digits) == 1L) {
    digits <- rep(digits, ncol(x))
  } else {
    digits <- rep_len(digits, ncol(x))
  }

  for (j in seq_along(x)) {
    if (is.numeric(x[[j]])) {
      x[[j]] <- ifelse(
        is.na(x[[j]]),
        "",
        formatC(x[[j]], digits = digits[j], format = "f")
      )
    } else {
      x[[j]] <- ifelse(is.na(x[[j]]), "", as.character(x[[j]]))
    }
  }

  align <- c("l", rep("r", max(ncol(x) - 1L, 0L)))

  out <- knitr::kable(
    x,
    caption = caption,
    col.names = col_names,
    digits = digits,
    align = align,
    format = if (knitr::is_html_output()) "html" else "pipe",
    table.attr = if (knitr::is_html_output()) {
      'class="simulation-preview-table"'
    } else {
      NULL
    }
  )

  if (knitr::is_html_output()) {
    out <- paste0(
      "<div class=\"simulation-table-wrap\">",
      out,
      "</div>"
    )
  }

  knitr::asis_output(out)
}

build_performance_table <- function(evaluation) {
  weighted <- evaluation$weighted

  data.frame(
    Metric = c(
      "Precision",
      "Recall",
      "F1 score",
      "Specificity",
      "False positive rate",
      "Balanced accuracy"
    ),
    Strict = c(
      evaluation$strict$precision,
      evaluation$strict$recall,
      evaluation$strict$f1,
      evaluation$strict$specificity,
      evaluation$strict$fpr,
      evaluation$strict$balanced_accuracy
    ),
    `Strict wt` = c(
      if (!is.null(weighted)) weighted$strict$precision else NA_real_,
      if (!is.null(weighted)) weighted$strict$recall else NA_real_,
      if (!is.null(weighted)) weighted$strict$f1 else NA_real_,
      NA_real_,
      NA_real_,
      NA_real_
    ),
    Fuzzy = c(
      evaluation$fuzzy$precision,
      evaluation$fuzzy$recall,
      evaluation$fuzzy$f1,
      evaluation$fuzzy$specificity,
      evaluation$fuzzy$fpr,
      evaluation$fuzzy$balanced_accuracy
    ),
    `Fuzzy wt` = c(
      if (!is.null(weighted)) weighted$fuzzy$precision else NA_real_,
      if (!is.null(weighted)) weighted$fuzzy$recall else NA_real_,
      if (!is.null(weighted)) weighted$fuzzy$f1 else NA_real_,
      NA_real_,
      NA_real_,
      NA_real_
    ),
    check.names = FALSE
  )
}

build_null_summary_table <- function(studies) {
  rows <- lapply(names(studies), function(ic_name) {
    study <- studies[[ic_name]]
    per_rep <- study$per_replicate
    summary <- study$study_summary

    mean_fp <- if (!is.null(summary$mean_false_positive_rate)) {
      summary$mean_false_positive_rate
    } else if (!is.null(per_rep$false_positive_rate)) {
      mean(per_rep$false_positive_rate, na.rm = TRUE)
    } else {
      NA_real_
    }

    median_fp <- if (!is.null(summary$median_false_positive_rate)) {
      summary$median_false_positive_rate
    } else if (!is.null(per_rep$false_positive_rate)) {
      stats::median(per_rep$false_positive_rate, na.rm = TRUE)
    } else {
      NA_real_
    }

    any_fp <- if (!is.null(summary$fraction_any_false_positive)) {
      summary$fraction_any_false_positive
    } else if (!is.null(per_rep$n_inferred_shifts)) {
      mean(per_rep$n_inferred_shifts > 0, na.rm = TRUE)
    } else {
      NA_real_
    }

    mean_inferred <- if (!is.null(summary$mean_inferred_shifts)) {
      summary$mean_inferred_shifts
    } else if (!is.null(per_rep$n_inferred_shifts)) {
      mean(per_rep$n_inferred_shifts, na.rm = TRUE)
    } else {
      NA_real_
    }

    data.frame(
      IC = ic_name,
      `Mean FP` = mean_fp,
      `Median FP` = median_fp,
      `Any FP` = any_fp,
      `Mean shifts` = mean_inferred,
      `Evaluable` = summary$n_evaluable_replicates,
      check.names = FALSE
    )
  })

  do.call(rbind, rows)
}

compute_effective_dimensionality <- function(sigma) {
  eigenvalues <- eigen(sigma, symmetric = TRUE, only.values = TRUE)$values
  weights <- eigenvalues / sum(eigenvalues)
  weights <- weights[weights > 0]
  exp(-sum(weights * log(weights)))
}

simulate_demo_ancestral_vcv <- function(
  n_traits = 10,
  var_mean = 1,
  var_sd = 0.1,
  cov_mean = 0.5,
  cov_sd = 0.25
) {
  variances <- abs(stats::rnorm(n_traits, var_mean, var_sd))
  lower_tri <- matrix(stats::rnorm(n_traits^2, cov_mean, cov_sd), nrow = n_traits)
  lower_tri[upper.tri(lower_tri)] <- 0
  sigma <- t(lower_tri) %*% lower_tri
  diag(sigma) <- diag(sigma) + variances
  sigma
}

rescale_demo_correlation <- function(sigma, scale_factor) {
  cor_mat <- stats::cov2cor(sigma)
  scaled_cor <- cor_mat
  scaled_cor[lower.tri(scaled_cor)] <- cor_mat[lower.tri(cor_mat)] * scale_factor
  scaled_cor[upper.tri(scaled_cor)] <- t(scaled_cor)[upper.tri(scaled_cor)]

  sd_vec <- sqrt(diag(sigma))
  new_vars <- (sd_vec^2) / scale_factor
  rescaled_sigma <- diag(sqrt(new_vars)) %*% scaled_cor %*% diag(sqrt(new_vars))

  list(
    cor_mat = cor_mat,
    scaled_cor = scaled_cor,
    new_vars = new_vars,
    rescaled_sigma = rescaled_sigma
  )
}

build_simulation_scenario_figure <- function(
  seed = 123,
  n_traits = 10,
  var_mean = 1,
  var_sd = 0.1,
  cov_mean = 0.5,
  cov_sd = 0.25,
  scale_factor = 0.5
) {
  if (!requireNamespace("ggplot2", quietly = TRUE) ||
      !requireNamespace("reshape2", quietly = TRUE) ||
      !requireNamespace("viridis", quietly = TRUE) ||
      !requireNamespace("patchwork", quietly = TRUE)) {
    stop("Rendering the simulation-scenarios figure requires ggplot2, reshape2, viridis, and patchwork.")
  }

  set.seed(seed)
  ancestral_sigma <- simulate_demo_ancestral_vcv(
    n_traits = n_traits,
    var_mean = var_mean,
    var_sd = var_sd,
    cov_mean = cov_mean,
    cov_sd = cov_sd
  )
  correlation_rescaled <- rescale_demo_correlation(ancestral_sigma, scale_factor)
  proportional_sigma <- scale_factor * ancestral_sigma

  vcv_list <- list(
    "Ancestral VCV" = ancestral_sigma,
    "Proportional VCV" = proportional_sigma,
    "Reduced Correlation VCV" = correlation_rescaled$rescaled_sigma
  )
  cor_list <- list(
    "Correlation Matrix (Anc)" = stats::cov2cor(ancestral_sigma),
    "Correlation Matrix (Prop)" = stats::cov2cor(proportional_sigma),
    "Correlation Matrix (Reduced)" = stats::cov2cor(correlation_rescaled$rescaled_sigma)
  )

  all_vals <- unlist(lapply(vcv_list, function(mat) mat))
  vcv_limits <- c(min(all_vals), max(all_vals))

  melt_matrix <- function(mat) {
    df <- reshape2::melt(mat)
    colnames(df) <- c("Trait1", "Trait2", "Value")
    df
  }

  clean_heatmap_theme <- ggplot2::theme(
    panel.grid = ggplot2::element_blank(),
    panel.background = ggplot2::element_blank(),
    panel.border = ggplot2::element_blank(),
    axis.text = ggplot2::element_blank(),
    axis.ticks = ggplot2::element_blank(),
    plot.title = ggplot2::element_text(size = 12, face = "bold"),
    legend.title = ggplot2::element_text(size = 11),
    legend.text = ggplot2::element_text(size = 10)
  )

  l_shaped_theme <- ggplot2::theme(
    panel.grid = ggplot2::element_blank(),
    panel.background = ggplot2::element_blank(),
    panel.border = ggplot2::element_blank(),
    axis.line = ggplot2::element_line(),
    axis.ticks = ggplot2::element_line(),
    axis.text.x = ggplot2::element_text(size = 10),
    axis.text.y = ggplot2::element_text(size = 10),
    axis.title = ggplot2::element_text(size = 11),
    plot.title = ggplot2::element_text(size = 12, face = "bold"),
    axis.ticks.length = grid::unit(0.25, "cm"),
    legend.title = ggplot2::element_text(size = 11),
    legend.text = ggplot2::element_text(size = 10)
  )

  vcv_plots <- lapply(names(vcv_list), function(name) {
    ggplot2::ggplot(melt_matrix(vcv_list[[name]]), ggplot2::aes(Trait1, Trait2, fill = Value)) +
      ggplot2::geom_tile() +
      viridis::scale_fill_viridis(
        name = "VCV Value",
        option = "D",
        limits = vcv_limits
      ) +
      ggplot2::labs(title = name, x = NULL, y = NULL) +
      ggplot2::theme_minimal(base_size = 11) +
      clean_heatmap_theme
  })

  cor_plots <- lapply(names(cor_list), function(name) {
    ggplot2::ggplot(melt_matrix(cor_list[[name]]), ggplot2::aes(Trait1, Trait2, fill = Value)) +
      ggplot2::geom_tile() +
      viridis::scale_fill_viridis(
        name = "Correlation",
        option = "B",
        limits = c(0, 0.75)
      ) +
      ggplot2::labs(title = name, x = NULL, y = NULL) +
      ggplot2::theme_minimal(base_size = 11) +
      clean_heatmap_theme
  })

  eigen_df <- do.call(rbind, list(
    data.frame(
      Eigenvalue = eigen(ancestral_sigma)$values,
      Rank = seq_len(n_traits),
      Scenario = "Ancestral"
    ),
    data.frame(
      Eigenvalue = eigen(correlation_rescaled$rescaled_sigma)$values,
      Rank = seq_len(n_traits),
      Scenario = "Correlation-scaled"
    ),
    data.frame(
      Eigenvalue = eigen(proportional_sigma)$values,
      Rank = seq_len(n_traits),
      Scenario = "Proportional-scaled"
    )
  ))

  eff_dim_df <- data.frame(
    Scenario = c("Ancestral", "Correlation-scaled", "Proportional-scaled"),
    EffDimensionality = c(
      compute_effective_dimensionality(ancestral_sigma),
      compute_effective_dimensionality(correlation_rescaled$rescaled_sigma),
      compute_effective_dimensionality(proportional_sigma)
    )
  )

  mean_var_df <- data.frame(
    Scenario = c("Ancestral", "Correlation-scaled", "Proportional-scaled"),
    MeanVariance = c(
      mean(diag(ancestral_sigma)),
      mean(diag(correlation_rescaled$rescaled_sigma)),
      mean(diag(proportional_sigma))
    )
  )

  p_effdim <- ggplot2::ggplot(eff_dim_df, ggplot2::aes(x = Scenario, y = EffDimensionality, fill = Scenario)) +
    ggplot2::geom_bar(stat = "identity") +
    ggplot2::labs(title = "Effective Dimensionality", y = "Effective Dimensionality", x = NULL) +
    ggplot2::theme_minimal(base_size = 11) +
    ggplot2::scale_fill_viridis_d() +
    l_shaped_theme +
    ggplot2::theme(
      axis.title.x = ggplot2::element_blank(),
      axis.text.x = ggplot2::element_blank(),
      legend.position = "none"
    )

  p_eig <- ggplot2::ggplot(eigen_df, ggplot2::aes(x = Rank, y = Eigenvalue, color = Scenario)) +
    ggplot2::geom_line(linewidth = 1.2) +
    ggplot2::labs(title = "Eigenvalue Spectra", x = "Rank", y = "Eigenvalue") +
    ggplot2::theme_minimal(base_size = 11) +
    ggplot2::scale_color_viridis_d() +
    l_shaped_theme +
    ggplot2::theme(legend.position = "none")

  p_trace <- ggplot2::ggplot(mean_var_df, ggplot2::aes(x = Scenario, y = MeanVariance, fill = Scenario)) +
    ggplot2::geom_bar(stat = "identity") +
    ggplot2::labs(title = "Mean Diagonal Variance", y = "Mean Variance", x = NULL) +
    ggplot2::theme_minimal(base_size = 11) +
    ggplot2::scale_fill_viridis_d() +
    l_shaped_theme +
    ggplot2::theme(
      axis.title.x = ggplot2::element_blank(),
      axis.text.x = ggplot2::element_blank()
    )

  vcv_plots[[1]] <- vcv_plots[[1]] + ggplot2::theme(legend.position = "none")
  vcv_plots[[2]] <- vcv_plots[[2]] + ggplot2::theme(legend.position = "none")
  cor_plots[[1]] <- cor_plots[[1]] + ggplot2::theme(legend.position = "none")
  cor_plots[[2]] <- cor_plots[[2]] + ggplot2::theme(legend.position = "none")

  title_text <- "Covariance Structure Under Three Simulation Scenarios"

  ((vcv_plots[[1]] + vcv_plots[[2]] + vcv_plots[[3]]) /
     (cor_plots[[1]] + cor_plots[[2]] + cor_plots[[3]]) /
     (p_effdim + p_eig + p_trace)) +
    patchwork::plot_annotation(
      title = title_text,
      tag_levels = "A",
      theme = ggplot2::theme(
        plot.title = ggplot2::element_text(size = 14, face = "bold", hjust = 0.5),
        plot.tag = ggplot2::element_text(size = 13, face = "bold")
      )
    )
}

pkg_file <- function(...) {
  candidates <- c(
    file.path("inst", ...),
    file.path("..", "inst", ...)
  )

  for (path in candidates) {
    if (file.exists(path)) {
      return(path)
    }
  }

  installed_path <- system.file(..., package = "bifrost")
  if (nzchar(installed_path)) {
    return(installed_path)
  }

  stop("Could not locate: ", file.path(...))
}
```

```{r, echo = FALSE, results = "asis"}
if (knitr::is_html_output()) cat("
<style>
img, .figure img, .figure > p > img {
  border: 0 !important;
  box-shadow: none !important;
  outline: none !important;
  background: transparent !important;
}

p.caption, .caption, table > caption {
  border-left: 0.32rem solid #d8dde3;
  color: #2f3439;
  font-size: inherit;
  line-height: inherit;
  margin: 0 3.8% 1.05rem 3.8%;
  padding: 0 0 0 1rem;
  text-align: left !important;
}

p.caption strong, .caption strong {
  font-weight: 700;
}

table {
  width: 100%;
}

table th, table td {
  vertical-align: top;
}

table code {
  white-space: nowrap;
}

.simulation-table-wrap {
  margin: 1em 0 1.15em 0;
  max-width: 100%;
  overflow-x: auto;
}

.simulation-table-wrap table {
  margin-bottom: 0;
}

.simulation-preview-table thead th {
  white-space: nowrap;
}
</style>
")
```

## Why use dataset-matched simulations

Before fitting an empirical `bifrost` search, it helps to know how the method behaves on datasets that look like the one you actually care about. A practical way to answer that question is to match the simulations to the empirical tree, trait dimensionality, and covariance scale, then ask how the search behaves under three scenarios:

1. a null model with no shifts, to estimate expected false positives,
2. the current generating model, where shifts are proportional covariance scalings, and
3. a deliberately non-generating robustness scenario, where trait correlations evolve in ways that violate strict proportional scaling.

This strategy follows the general logic of the passerine body-plan study of Berv et al. (2026), but it is useful well beyond that one application. Fit a dataset-matched template, simulate the three scenario classes, use those results to choose search settings, and then carry the selected search back into the empirical analysis. On large trees, repeated subtree resampling is especially helpful because it lets simulations vary across many local topologies and parameter draws without requiring a full-tree search on every replicate.

A typical session starts by loading `bifrost` and `ape`.

```{r load_packages, eval = FALSE}
library(bifrost)
library(ape)
```

## Build a dataset-matched template

The packaged passerine skeleton example provides a useful calibration starting point: 12 skeletal traits are modeled jointly while log body mass is kept as a phylogenetic covariate.

```{r load_bodyplan_example}
tree_path <- pkg_file("extdata", "avian-skeleton", "passerine_bodyplan_tree.tre")
trait_path <- pkg_file("extdata", "avian-skeleton", "passerine_bodyplan_data.RDS")

bird_tree <- ape::read.tree(tree_path)
bodyplan_data <- readRDS(trait_path)
bodyplan_data <- as.matrix(bodyplan_data[bird_tree$tip.label, , drop = FALSE])

skeletal_cols <- setdiff(colnames(bodyplan_data), "vertnet_mass")
bodyplan_data <- bodyplan_data[, c(skeletal_cols, "vertnet_mass"), drop = FALSE]

formula_str <- sprintf(
  "trait_data[, 1:%d] ~ trait_data[, %d]",
  length(skeletal_cols),
  ncol(bodyplan_data)
)

c(
  species = nrow(bodyplan_data),
  skeletal_traits = length(skeletal_cols),
  total_columns = ncol(bodyplan_data)
)
```

The template comes from a single-regime global fit. That fit provides the empirical variance and covariance scale used to generate a fresh ancestral covariance matrix for each simulation replicate. Under the hood, it uses the penalized high-dimensional phylogenetic GLS framework of [Clavel, Aristide, and Morlon (2019)](https://doi.org/10.1093/sysbio/syy045), as implemented in `mvMORPH` ([Clavel, Escarguel, and Merceron 2015](https://doi.org/10.1111/2041-210X.12420)).

```{r create_template_code, eval = FALSE}
bodyplan_template <- createSimulationTemplate(
  baseline_tree = bird_tree,
  trait_data = bodyplan_data,
  formula = formula_str,
  response_columns = seq_along(skeletal_cols),
  predictor_columns = ncol(bodyplan_data),
  method = "LL",
  error = TRUE
)
```

```{r create_template_live, echo = FALSE, message = FALSE, warning = FALSE, cache = TRUE}
bodyplan_template_live <- createSimulationTemplate(
  baseline_tree = bird_tree,
  trait_data = bodyplan_data,
  formula = formula_str,
  response_columns = seq_along(skeletal_cols),
  predictor_columns = ncol(bodyplan_data),
  method = "LL",
  error = TRUE
)
```

```{r create_template_print, echo = FALSE}
bodyplan_template_live
```

The template stores the aligned calibration data, the fitted global `mvgls` model, and the empirical variance/covariance summaries used to generate a fresh ancestral covariance matrix for each simulation replicate.

Conceptually, that calibration step is broader than the specific formula examples shown here. The simulation workflow operates on residual variation from a single global fit: it simulates a new response block around the fitted mean structure, then evaluates intercept-only shift searches on that regenerated response block. In that sense, the underlying logic can accommodate much richer empirical mean structures than the legacy indexed examples might suggest. The main practical requirement is that the response columns remain raw numeric variables that can be regenerated in each replicate.

## When an intercept-only shortcut helps

The passerine body-plan study used one additional shortcut that is worth making explicit. Body mass was included when estimating the global covariance scale, but the simulation searches themselves were run as intercept-only searches. In practical terms, the calibration fit captures the mean and residual structure of the empirical dataset once, and the simulation studies then ask how the shift search behaves on the regenerated multivariate response block.

Put differently, the simulation engine is targeting the residual covariance structure implied by the global model fit, not just one particular formula syntax. The intercept-only shortcut simply strips that idea down to the multivariate response block alone.

The same residual-focused approximation can also be reproduced in `bifrost` with an intercept-only, response-only template when no calibration covariates are needed:

```{r manuscript_style_template, eval = FALSE}
bodyplan_response_only <- bodyplan_data[, skeletal_cols, drop = FALSE]

bodyplan_template_intercept <- createSimulationTemplate(
  baseline_tree = bird_tree,
  trait_data = bodyplan_response_only,
  formula = "trait_data ~ 1",
  method = "LL",
  error = TRUE
)
```

Using `bodyplan_template_intercept` makes the calibration step target the multivariate response block alone. That is the simplest special case. More generally, a richer global template such as `bodyplan_template` can still feed the same intercept-only simulation studies.

The same shortcut also works with subset-response intercept formulas such as `"trait_data[, 1:12] ~ 1"`, as long as `response_columns` identifies the columns to simulate.

## Simulation scenarios

The figure below summarizes the three simulation scenarios used throughout this vignette. It reproduces the Supplementary Figure 1 demonstration from Berv et al. (2026): a single ancestral covariance matrix, a proportional rescaling, and a reduced-correlation transform. The goal here is to show exactly what changes under each scenario before those same ideas are applied to dataset-matched subtree resampling.

```{r simulation_scenarios_figure, echo = FALSE, dev = "svglite", fig.ext = "svg", fig.width = 8.8, fig.height = 8.4, out.width = "100%", fig.cap = "Simulation scenarios reproduced from the Supplementary Figure 1 workflow in Berv et al. (2026). Panels A-C show the ancestral VCV, a proportional rescaling with s = 0.5, and a reduced-correlation scenario in which off-diagonal correlations are scaled by s = 0.5 while variances are adjusted to preserve comparable covariances. Panels D-F show the corresponding correlation matrices. Panels G-I summarize effective dimensionality, eigenvalue spectra, and mean diagonal variance. For a fuller discussion of the panel logic, see Supplementary Figure 1 in Berv et al. (2026)."}
build_simulation_scenario_figure(
  seed = 123,
  n_traits = 10,
  var_mean = 1,
  var_sd = 0.1,
  cov_mean = 0.5,
  cov_sd = 0.25,
  scale_factor = 0.5
)
```

The main contrast is between proportional scaling and reduced-correlation scaling. Proportional scaling changes overall magnitude while leaving the correlation pattern intact, whereas the reduced-correlation scenario weakens trait associations and compensates with larger variances. The passerine examples below use these same two simulation styles under repeated subtree resampling from the full tree.

```{r preview_template_setup, echo = FALSE, message = FALSE, warning = FALSE, cache = TRUE}
preview_template <- bodyplan_template_live
```

## Calibration guardrails

Keep these guardrails in mind before launching a large study.

- Once the template has been fitted, the simulation studies operate on the regenerated response block alone. Predictor terms matter in the global calibration fit, not in each subtree-resampled search replicate.

- Reproducibility for replicated studies is controlled by the wrapper-level `seed` argument in `runFalsePositiveSimulationStudy()` and `runShiftRecoverySimulationStudy()`. There is no need to pass a separate seed inside `simulation_options`; the wrappers manage replicate-level seeding internally.

- The study wrappers inherit key fit settings such as `method` and `error` from the simulation template. In many cases you can leave those out of `search_options` unless you intentionally want to override the template defaults.

- Inspect `per_replicate$n_candidates` after a run. If `min_descendant_tips` is set so high that very few candidate shifts remain, the study will be much less informative about expected model performance.

- Treat a replicate with `n_candidates == 0` as unevaluable rather than as evidence for good recovery performance. In that case, recovery specificity, false-positive rate, and balanced accuracy are reported as `NA`, and the usual fix is to lower `min_descendant_tips` until a meaningful candidate set is available.

## 1. Inspect One Fixed Search

This worked example keeps the search settings fixed and asks a simple question: what does expected performance look like on repeated 200-tip subtree resamples from the passerine tree? The template is still calibrated from the global body-mass model above, but the simulated searches are intercept-only. Holding the settings fixed makes the null, proportional, and correlation scenarios easier to interpret before comparing alternative thresholds or clade-size filters.

The search is intentionally conservative: `shift_acceptance_threshold = 10`, `min_descendant_tips = 10`, and five simulated shifts ranging from 10 to 20 descendant tips. These settings are not being optimized yet. They provide a stable baseline for understanding how null and shifted datasets behave.

```{r medium_case_study_setup, echo = FALSE, message = FALSE, warning = FALSE, cache = TRUE}
if (requireNamespace("RhpcBLASctl", quietly = TRUE)) {
  RhpcBLASctl::blas_set_num_threads(1)
  RhpcBLASctl::omp_set_num_threads(1)
}

medium_case_template <- preview_template

medium_case_simulation_options <- list(
  num_shifts = 5,
  min_shift_tips = 10,
  max_shift_tips = 20,
  scale_factor_range = c(0.1, 2.0),
  exclude_range = c(0.5, 1.5),
  buffer = 3
)

medium_case_search <- function(IC) {
  list(
    formula = "trait_data ~ 1",
    IC = IC,
    shift_acceptance_threshold = 10,
    min_descendant_tips = 10,
    method = "LL",
    error = TRUE,
    num_cores = 1,
    uncertaintyweights_par = TRUE,
    plot = FALSE,
    store_model_fit_history = FALSE
  )
}

gic_case_null <- runFalsePositiveSimulationStudy(
  template = medium_case_template,
  n_replicates = 20,
  tree_tip_count = 200,
  search_options = medium_case_search("GIC"),
  num_cores = 8,
  seed = 11
)
gic_case_prop <- runShiftRecoverySimulationStudy(
  template = medium_case_template,
  n_replicates = 20,
  tree_tip_count = 200,
  simulation_options = utils::modifyList(
    medium_case_simulation_options,
    list(scale_mode = "proportional")
  ),
  search_options = medium_case_search("GIC"),
  fuzzy_distance = 2,
  weighted = TRUE,
  num_cores = 8,
  seed = 1011
)
gic_case_corr <- runShiftRecoverySimulationStudy(
  template = medium_case_template,
  n_replicates = 20,
  tree_tip_count = 200,
  simulation_options = utils::modifyList(
    medium_case_simulation_options,
    list(scale_mode = "correlation")
  ),
  search_options = medium_case_search("GIC"),
  fuzzy_distance = 2,
  weighted = TRUE,
  num_cores = 8,
  seed = 2011
)

bic_case_null <- runFalsePositiveSimulationStudy(
  template = medium_case_template,
  n_replicates = 20,
  tree_tip_count = 200,
  search_options = medium_case_search("BIC"),
  num_cores = 8,
  seed = 22
)
bic_case_prop <- runShiftRecoverySimulationStudy(
  template = medium_case_template,
  n_replicates = 20,
  tree_tip_count = 200,
  simulation_options = utils::modifyList(
    medium_case_simulation_options,
    list(scale_mode = "proportional")
  ),
  search_options = medium_case_search("BIC"),
  fuzzy_distance = 2,
  weighted = TRUE,
  num_cores = 8,
  seed = 1022
)
bic_case_corr <- runShiftRecoverySimulationStudy(
  template = medium_case_template,
  n_replicates = 20,
  tree_tip_count = 200,
  simulation_options = utils::modifyList(
    medium_case_simulation_options,
    list(scale_mode = "correlation")
  ),
  search_options = medium_case_search("BIC"),
  fuzzy_distance = 2,
  weighted = TRUE,
  num_cores = 8,
  seed = 2022
)
```

Blank cells in the weighted columns for specificity, false-positive rate, and balanced accuracy are expected. Weighted summaries are defined only for precision, recall, and F1 because the weights apply to inferred shifts rather than to true negatives.

```{r medium_case_study_tables, echo = FALSE, results = "asis"}
render_preview_table(
  build_null_summary_table(list(GIC = gic_case_null, BIC = bic_case_null)),
  caption = "Fixed-setting case study with 20 repeated 200-tip subsamples. In this preview, both GIC and BIC remain fully conservative under the null, with no inferred false positives.",
  digits = c(0, 3, 3, 3, 3, 0)
)

render_preview_table(
  build_performance_table(gic_case_prop$evaluation),
  caption = "Fixed-setting case study with 20 repeated 200-tip subsamples. Under proportional scaling, GIC recovers shifts well, especially under fuzzy matching.",
  digits = c(0, 3, 3, 3, 3)
)

render_preview_table(
  build_performance_table(bic_case_prop$evaluation),
  caption = "Fixed-setting case study with 20 repeated 200-tip subsamples. Under proportional scaling, BIC stays a bit more conservative while retaining strong recovery.",
  digits = c(0, 3, 3, 3, 3)
)

render_preview_table(
  build_performance_table(gic_case_corr$evaluation),
  caption = "Fixed-setting case study with 20 repeated 200-tip subsamples. Under correlation change, GIC still recovers many shifts, with fuzzy matching outperforming strict matching.",
  digits = c(0, 3, 3, 3, 3)
)

render_preview_table(
  build_performance_table(bic_case_corr$evaluation),
  caption = "Fixed-setting case study with 20 repeated 200-tip subsamples. Under correlation change, BIC becomes more conservative and recovery drops most clearly under strict matching.",
  digits = c(0, 3, 3, 3, 3)
)
```

Taken together, these tables show how one fixed search behaves under repeated subtree resampling: how often it infers shifts under the null, how well it recovers shifts when the generating model is proportional, and how much that recovery changes when trait correlations evolve in a non-proportional way.

That kind of case study is useful when the goal is to understand the implications of one specific intercept-only search. When the goal is calibration, the next step is to estimate false positives, recovery, and robustness more systematically, then compare plausible thresholds and clade-size filters in a small grid.

## 2. Estimate False Positives

The null study asks the most conservative performance question: if the real dataset contained no shifts at all, how often would the search still infer one? The template below is still calibrated from the global body-mass model, but the simulation search itself remains intercept-only.

```{r false_positive_study, eval = FALSE}
fp_study <- runFalsePositiveSimulationStudy(
  template = bodyplan_template,
  n_replicates = 100,
  tree_tip_count = 250,
  simulation_options = list(),
  search_options = list(
    formula = "trait_data ~ 1",
    min_descendant_tips = 10,
    shift_acceptance_threshold = 10,
    IC = "GIC",
    num_cores = 1,
    method = "LL",
    error = TRUE,
    plot = FALSE,
    store_model_fit_history = FALSE
  ),
  num_cores = 8,
  seed = 5
)
```

```{r false_positive_inspect, eval = FALSE}
fp_study
head(fp_study$per_replicate)
fp_study$study_summary
```

Null simulations are especially useful for screening conservative choices such as `shift_acceptance_threshold`, `IC`, and `min_descendant_tips`.

## 3. Estimate Recovery Under Proportional Scaling

Recovery under proportional scaling evaluates the scenario that matches the current `bifrost` BMM search assumptions: shifted covariance matrices are scalar multiples of the background covariance matrix. As above, the template may be calibrated from a richer global model, but the simulation search itself remains intercept-only.

```{r proportional_recovery_study, eval = FALSE}
recovery_prop <- runShiftRecoverySimulationStudy(
  template = bodyplan_template,
  n_replicates = 100,
  tree_tip_count = 250,
  simulation_options = list(
    num_shifts = 5,
    min_shift_tips = 10,
    max_shift_tips = 20,
    scale_mode = "proportional",
    scale_factor_range = c(0.1, 2.0),
    exclude_range = c(0.5, 1.5),
    buffer = 3
  ),
  search_options = list(
    formula = "trait_data ~ 1",
    min_descendant_tips = 10,
    shift_acceptance_threshold = 10,
    IC = "GIC",
    num_cores = 1,
    method = "LL",
    error = TRUE,
    uncertaintyweights_par = TRUE,
    plot = FALSE,
    store_model_fit_history = FALSE
  ),
  fuzzy_distance = 2,
  weighted = TRUE,
  num_cores = 8,
  seed = 5
)
```

```{r proportional_recovery_inspect, eval = FALSE}
recovery_prop
recovery_prop$study_summary
recovery_prop$evaluation$strict
recovery_prop$evaluation$fuzzy
recovery_prop$evaluation$weighted
```

## 4. Check Robustness Under Correlation Change

It is also useful to test a deliberately mis-specified scenario in which trait correlations evolve. This asks how well the same intercept-only search settings behave when the data depart from strict proportional VCV scaling.

```{r correlation_recovery_study, eval = FALSE}
recovery_corr <- runShiftRecoverySimulationStudy(
  template = bodyplan_template,
  n_replicates = 100,
  tree_tip_count = 250,
  simulation_options = list(
    num_shifts = 5,
    min_shift_tips = 10,
    max_shift_tips = 20,
    scale_mode = "correlation",
    scale_factor_range = c(0.1, 2.0),
    exclude_range = c(0.5, 1.5),
    buffer = 3
  ),
  search_options = list(
    formula = "trait_data ~ 1",
    min_descendant_tips = 10,
    shift_acceptance_threshold = 10,
    IC = "GIC",
    num_cores = 1,
    method = "LL",
    error = TRUE,
    uncertaintyweights_par = TRUE,
    plot = FALSE,
    store_model_fit_history = FALSE
  ),
  fuzzy_distance = 2,
  weighted = TRUE,
  num_cores = 8,
  seed = 5
)
```

```{r correlation_recovery_inspect, eval = FALSE}
recovery_corr
recovery_corr$study_summary
recovery_corr$evaluation$fuzzy
```

This scenario is not the current generating model for `bifrost`, so the goal is not perfect recovery. Instead, it is a robustness check: do the same search settings still behave reasonably when the data depart from the assumed model in a biologically plausible way?

## 5. Choose Settings from a Search Grid

The fixed-setting case study describes how one search behaves. Choosing settings is a different task. A small simulation grid is usually more informative because it makes the tradeoff between `shift_acceptance_threshold` and `min_descendant_tips` visible instead of relying on a single ad hoc choice.

In practice, this usually means running one grid for GIC and a second grid for BIC. If the empirical analysis will be fitted with GIC, it makes the most sense to choose the best threshold and clade-size settings within the GIC results. The same logic applies to BIC. Pooling them into a single ranking is less useful because the two criteria penalize model complexity differently.

For a compact example, the grid varies three acceptance thresholds (`10`, `20`, `30`) and two minimum clade sizes (`10`, `20`). Five null and five recovery replicates per setting are enough to show the tuning logic, and the same grid scales directly to larger replicate counts.

```{r preview_tuning_setup, echo = FALSE, message = FALSE, warning = FALSE, cache = TRUE}
preview_grid_args <- list(
  shift_acceptance_thresholds = c(10, 20, 30),
  min_descendant_tips_values = c(10, 20),
  tree_tip_count = 200,
  null_replicates = 5,
  recovery_replicates = 5,
  proportional_simulation_options = list(
    num_shifts = 5,
    min_shift_tips = 10,
    max_shift_tips = 20,
    scale_factor_range = c(0.1, 2.0),
    exclude_range = c(0.5, 1.5),
    buffer = 3
  ),
  base_search_options = list(
    formula = "trait_data ~ 1",
    method = "LL",
    error = TRUE,
    num_cores = 1,
    uncertaintyweights_par = TRUE,
    plot = FALSE,
    store_model_fit_history = FALSE
  ),
  weighted = TRUE,
  num_cores = 8,
  seed = 5,
  store_studies = FALSE
)

preview_cols <- c(
  "shift_acceptance_threshold",
  "min_descendant_tips",
  "null_mean_false_positive_rate",
  "proportional_fuzzy_f1",
  "correlation_fuzzy_f1"
)
preview_labels <- c(
  "Threshold",
  "Min clade",
  "Null FP",
  "Prop. Fuzzy F1",
  "Corr. Fuzzy F1"
)

gic_preview_grid <- do.call(
  runSearchTuningGrid,
  c(list(template = preview_template, IC = "GIC"), preview_grid_args)
)
bic_preview_grid <- do.call(
  runSearchTuningGrid,
  c(list(template = preview_template, IC = "BIC"), preview_grid_args)
)

gic_preview_tuned <- selectTunedSearchParameters(
  gic_preview_grid,
  max_false_positive_rate = 0.10,
  max_any_false_positive = 0.50,
  min_evaluable_fraction = 0.50,
  primary_metric = "weighted_fuzzy_f1",
  tie_break = "conservative"
)
bic_preview_tuned <- selectTunedSearchParameters(
  bic_preview_grid,
  max_false_positive_rate = 0.10,
  max_any_false_positive = 0.50,
  min_evaluable_fraction = 0.50,
  primary_metric = "weighted_fuzzy_f1",
  tie_break = "conservative"
)

preview_recommendations <- rbind(
  data.frame(IC = "GIC", gic_preview_tuned$selected_row[, preview_cols], row.names = NULL),
  data.frame(IC = "BIC", bic_preview_tuned$selected_row[, preview_cols], row.names = NULL)
)
```

Each row represents one threshold/clade-size combination evaluated under repeated subtree resampling.

### GIC

```{r preview_tuning_gic_table, echo = FALSE, results = "asis"}
render_preview_table(
  gic_preview_grid$summary_table[, preview_cols],
  caption = "Preview tuning grid with 5 repeated 200-tip subsamples per setting. In this GIC preview, all settings keep null false positives at zero, and recovery is strongest at threshold 10 with a minimum clade size of 10.",
  col_names = preview_labels,
  digits = c(0, 0, 3, 3, 3)
)
```

### BIC

```{r preview_tuning_bic_table, echo = FALSE, results = "asis"}
render_preview_table(
  bic_preview_grid$summary_table[, preview_cols],
  caption = "Preview tuning grid with 5 repeated 200-tip subsamples per setting. In this BIC preview, all settings also remain conservative under the null, with the strongest recovery at threshold 10 and a minimum clade size of 10.",
  col_names = preview_labels,
  digits = c(0, 0, 3, 3, 3)
)
```

### Recommended settings

```{r preview_tuning_selected_table, echo = FALSE, results = "asis"}
render_preview_table(
  preview_recommendations,
  caption = "Selected settings from the repeated-subsampling GIC and BIC preview grids. In this preview, both IC families choose threshold 10 and a minimum clade size of 10 because that combination preserves null conservatism while maximizing fuzzy recovery.",
  col_names = c("IC", preview_labels),
  digits = c(0, 0, 0, 3, 3, 3)
)
```

The selected rows follow a simple rule. First, discard settings with too many false positives or too few evaluable replicates. Among the settings that remain, prefer the one with the strongest weighted fuzzy recovery, and break close ties conservatively in favor of larger thresholds or clade sizes.

The same code scales directly to larger runs, first for GIC and then for BIC.

```{r tune_gic_grid, eval = FALSE}
gic_grid <- runSearchTuningGrid(
  template = bodyplan_template,
  IC = "GIC",
  shift_acceptance_thresholds = c(10, 20, 30),
  min_descendant_tips_values = c(10, 20),
  tree_tip_count = 250,
  null_replicates = 100,
  recovery_replicates = 100,
  proportional_simulation_options = list(
    num_shifts = 5,
    min_shift_tips = 10,
    max_shift_tips = 20,
    scale_factor_range = c(0.1, 2.0),
    exclude_range = c(0.5, 1.5),
    buffer = 3
  ),
  base_search_options = list(
    formula = "trait_data ~ 1",
    method = "LL",
    error = TRUE,
    num_cores = 1,
    uncertaintyweights_par = TRUE,
    plot = FALSE,
    store_model_fit_history = FALSE
  ),
  weighted = TRUE,
  num_cores = 8,
  seed = 5
)
```

```{r select_gic_grid, eval = FALSE}
gic_tuned <- selectTunedSearchParameters(
  gic_grid,
  max_false_positive_rate = 0.10,
  max_any_false_positive = 0.50,
  min_evaluable_fraction = 0.50,
  primary_metric = "weighted_fuzzy_f1",
  tie_break = "conservative"
)

gic_tuned$selected_row
gic_tuned$recommended_search_options
```

BIC follows the same pattern:

```{r tune_bic_grid, eval = FALSE}
bic_grid <- runSearchTuningGrid(
  template = bodyplan_template,
  IC = "BIC",
  shift_acceptance_thresholds = c(10, 20, 30),
  min_descendant_tips_values = c(10, 20),
  tree_tip_count = 250,
  null_replicates = 100,
  recovery_replicates = 100,
  proportional_simulation_options = list(
    num_shifts = 5,
    min_shift_tips = 10,
    max_shift_tips = 20,
    scale_factor_range = c(0.1, 2.0),
    exclude_range = c(0.5, 1.5),
    buffer = 3
  ),
  base_search_options = list(
    formula = "trait_data ~ 1",
    method = "LL",
    error = TRUE,
    num_cores = 1,
    uncertaintyweights_par = TRUE,
    plot = FALSE,
    store_model_fit_history = FALSE
  ),
  weighted = TRUE,
  num_cores = 8,
  seed = 5
)
```

```{r select_bic_grid, eval = FALSE}
bic_tuned <- selectTunedSearchParameters(
  bic_grid,
  max_false_positive_rate = 0.10,
  max_any_false_positive = 0.50,
  min_evaluable_fraction = 0.50,
  primary_metric = "weighted_fuzzy_f1",
  tie_break = "conservative"
)

bic_tuned$selected_row
bic_tuned$recommended_search_options
```

This yields two tuned search configurations:

1. the best threshold/clade-size combination within GIC, and
2. the best threshold/clade-size combination within BIC.

That is usually easier to interpret than one global optimization across both IC families.

## 6. Use Tuned Settings in the Empirical Search

The selected search options are ready to pass directly into `searchOptimalConfiguration()`. For example, if you decide to proceed with the GIC-tuned workflow:

```{r empirical_search_after_simulations, eval = FALSE}
empirical_result <- do.call(
  searchOptimalConfiguration,
  utils::modifyList(
    gic_tuned$recommended_search_options,
    list(
      baseline_tree = bird_tree,
      trait_data = bodyplan_data,
      num_cores = 20,
      store_model_fit_history = TRUE,
      verbose = TRUE
    )
  )
)
```

## Intercept-Only Data

This approach also applies to intercept-only datasets such as GPA-aligned landmark matrices. Here `fish_tree` and `fish_data` stand in for any aligned intercept-only dataset, including the jaw-shape example used in the companion vignette. In that case the template call is simpler because the multivariate response is just the full trait matrix:

```{r intercept_only_template, eval = FALSE}
jaw_template <- createSimulationTemplate(
  baseline_tree = fish_tree,
  trait_data = fish_data,
  formula = "trait_data ~ 1",
  method = "H&L",
  error = TRUE
)
```

The downstream null, proportional, and correlation-scenario studies are then identical in structure; only the template and search options change.

## Calibration Summary

The new simulation API makes this dataset-matched assessment strategy reusable:

1. fit a global empirical template,
2. estimate false positives under the null model,
3. measure recovery under the generating model,
4. test robustness under the non-generating correlation scenario, and
5. choose empirical search settings from fixed-IC tuning grids.

That is the recommended way to calibrate `bifrost` for a new dataset whose tree size, trait dimension, and predictor structure may differ substantially from the original passerine body-plan example.

## References

Berv, J. S., Probst, C. M., Claramunt, S., Shipley, J. R., Friedman, M., Smith, S. A., Fouhey, D. F., and Weeks, B. C. (2026). *Rates of passerine body plan evolution in time and space*. Nature Ecology & Evolution (in press).

Clavel, J., Aristide, L., and Morlon, H. (2019). *A Penalized Likelihood Framework for High-Dimensional Phylogenetic Comparative Methods and an Application to New-World Monkeys Brain Evolution*. [https://doi.org/10.1093/sysbio/syy045](https://doi.org/10.1093/sysbio/syy045)

Clavel, J., Escarguel, G., and Merceron, G. (2015). *mvMORPH: an R package for fitting multivariate evolutionary models to morphometric data*. *Methods in Ecology and Evolution*, 6(11), 1311-1319. [https://doi.org/10.1111/2041-210X.12420](https://doi.org/10.1111/2041-210X.12420)

Smith, S. A., Walker-Hale, N., and Parins-Fukuchi, C. T. (2023). *Compositional shifts associated with major evolutionary transitions in plants*. *New Phytologist*, 239(6), 2404-2415. [https://doi.org/10.1111/nph.19099](https://doi.org/10.1111/nph.19099)
