---
title: "Empirically Calibrated Simulations for bifrost, Part 1: Performance"
output:
  rmarkdown::html_vignette: default
  rmarkdown::pdf_document:
    latex_engine: xelatex
vignette: >
  %\VignetteIndexEntry{Empirically Calibrated Simulations for bifrost, Part 1: Performance}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  out.width = "100%",
  dpi = 170,
  fig.retina = 2,
  dev = "png"
)

vignette_figure_caption <- function(html, latex) {
  pandoc_to <- knitr::opts_knit$get("rmarkdown.pandoc.to")
  if (is.character(pandoc_to) && length(pandoc_to) == 1L && grepl("^html", pandoc_to)) html else latex
}

Sys.setenv(
  OMP_NUM_THREADS = "1",
  OPENBLAS_NUM_THREADS = "1",
  MKL_NUM_THREADS = "1",
  VECLIB_MAXIMUM_THREADS = "1",
  BLIS_NUM_THREADS = "1",
  NUMEXPR_NUM_THREADS = "1"
)

if (!exists("createSimulationTemplate", mode = "function")) {
  pkg_root <- if (file.exists("DESCRIPTION")) {
    "."
  } else if (file.exists(file.path("..", "DESCRIPTION"))) {
    ".."
  } else {
    NULL
  }

  if (!is.null(pkg_root) && requireNamespace("pkgload", quietly = TRUE)) {
    pkgload::load_all(pkg_root, quiet = TRUE, export_all = FALSE, helpers = FALSE)
  } else {
    library(bifrost)
  }
}

render_preview_table <- function(x, caption, col_names = colnames(x), digits = 3) {
  x <- as.data.frame(x, stringsAsFactors = FALSE, check.names = FALSE)
  rownames(x) <- NULL
  is_html <- knitr::is_html_output()
  is_latex <- knitr::is_latex_output()

  if (length(digits) == 1L) {
    digits <- rep(digits, ncol(x))
  } else {
    digits <- rep_len(digits, ncol(x))
  }

  for (j in seq_along(x)) {
    if (is.numeric(x[[j]])) {
      x[[j]] <- ifelse(
        is.na(x[[j]]),
        "",
        formatC(x[[j]], digits = digits[j], format = "f")
      )
    } else {
      x[[j]] <- ifelse(is.na(x[[j]]), "", as.character(x[[j]]))
    }
  }

  align <- c("l", rep("r", max(ncol(x) - 1L, 0L)))

  out <- knitr::kable(
    x,
    caption = if (is_latex) NULL else caption,
    col.names = col_names,
    digits = digits,
    align = align,
    format = if (is_html) "html" else if (is_latex) "latex" else "pipe",
    table.attr = if (is_html) {
      'class="simulation-preview-table"'
    } else {
      NULL
    }
  )

  if (is_html) {
    out <- sub(
      "(<caption>)(Table [0-9]+\\.)([[:space:]])",
      "\\1<strong>\\2</strong>\\3",
      out
    )
    out <- paste0(
      "<div class=\"simulation-table-wrap\">",
      out,
      "</div>"
    )
  } else if (is_latex) {
    caption_label <- sub("^(Table [0-9]+\\.).*", "\\1", caption)
    caption_body <- sub("^Table [0-9]+\\.[[:space:]]*", "", caption)
    out <- paste0(
      "\\noindent\\textbf{", caption_label, "} ", caption_body,
      "\n\n", out
    )
  }

  knitr::asis_output(out)
}

compute_effective_dimensionality <- function(sigma) {
  eigenvalues <- eigen(
    stats::cov2cor(sigma),
    symmetric = TRUE,
    only.values = TRUE
  )$values
  weights <- eigenvalues / sum(eigenvalues)
  weights <- weights[weights > 0]
  exp(-sum(weights * log(weights)))
}

simulate_demo_ancestral_vcv <- function(
  n_traits = 10,
  var_mean = 1,
  var_sd = 0.1,
  cov_mean = 0.5,
  cov_sd = 0.25
) {
  variances <- abs(stats::rnorm(n_traits, var_mean, var_sd))
  lower_tri <- matrix(stats::rnorm(n_traits^2, cov_mean, cov_sd), nrow = n_traits)
  lower_tri[upper.tri(lower_tri)] <- 0
  sigma <- t(lower_tri) %*% lower_tri
  diag(sigma) <- diag(sigma) + variances
  sigma
}

rescale_demo_integration_rate <- function(sigma, integration_power) {
  cor_mat <- stats::cov2cor(sigma)
  eig <- eigen(cor_mat, symmetric = TRUE)
  powered_cor <- eig$vectors %*%
    diag(pmax(eig$values, 1e-8)^integration_power, nrow = nrow(sigma)) %*%
    t(eig$vectors)
  powered_cor <- stats::cov2cor(powered_cor)
  sd_vec <- sqrt(diag(sigma))
  fixed_marginal_sigma <- diag(sd_vec) %*% powered_cor %*% diag(sd_vec)
  variance_scale <- 1 / integration_power
  rescaled_sigma <- variance_scale * fixed_marginal_sigma

  list(
    cor_mat = cor_mat,
    powered_cor = powered_cor,
    variance_scale = variance_scale,
    rescaled_sigma = rescaled_sigma
  )
}

build_simulation_scenario_figure <- function(
  seed = 123,
  n_traits = 10,
  var_mean = 1,
  var_sd = 0.1,
  cov_mean = 0.5,
  cov_sd = 0.25,
  proportional_scale = 0.5,
  integration_power = 0.65
) {
  if (!requireNamespace("ggplot2", quietly = TRUE) ||
      !requireNamespace("viridis", quietly = TRUE) ||
      !requireNamespace("patchwork", quietly = TRUE)) {
    stop("Rendering the simulation-scenarios figure requires ggplot2, viridis, and patchwork.")
  }

  set.seed(seed)
  ancestral_sigma <- simulate_demo_ancestral_vcv(
    n_traits = n_traits,
    var_mean = var_mean,
    var_sd = var_sd,
    cov_mean = cov_mean,
    cov_sd = cov_sd
  )
  integration_rescaled <- rescale_demo_integration_rate(
    ancestral_sigma,
    integration_power
  )
  proportional_sigma <- proportional_scale * ancestral_sigma

  vcv_list <- list(
    "Ancestral VCV" = ancestral_sigma,
    "Proportional VCV" = proportional_sigma,
    "Reduced Integration / Faster Rate" = integration_rescaled$rescaled_sigma
  )
  cor_list <- list(
    "Correlation Matrix (Anc)" = stats::cov2cor(ancestral_sigma),
    "Correlation Matrix (Prop)" = stats::cov2cor(proportional_sigma),
    "Correlation Matrix (Reduced Integration)" =
      stats::cov2cor(integration_rescaled$rescaled_sigma)
  )

  all_vals <- unlist(lapply(vcv_list, function(mat) mat))
  vcv_limits <- c(min(all_vals), max(all_vals))

  melt_matrix <- function(mat) {
    df <- as.data.frame(as.table(mat), stringsAsFactors = FALSE)
    names(df) <- c("Trait1", "Trait2", "Value")
    df
  }

  clean_heatmap_theme <- ggplot2::theme(
    panel.grid = ggplot2::element_blank(),
    panel.background = ggplot2::element_blank(),
    panel.border = ggplot2::element_blank(),
    axis.text = ggplot2::element_blank(),
    axis.ticks = ggplot2::element_blank(),
    plot.title = ggplot2::element_text(size = 12, face = "bold"),
    legend.title = ggplot2::element_text(size = 11),
    legend.text = ggplot2::element_text(size = 10)
  )

  l_shaped_theme <- ggplot2::theme(
    panel.grid = ggplot2::element_blank(),
    panel.background = ggplot2::element_blank(),
    panel.border = ggplot2::element_blank(),
    axis.line = ggplot2::element_line(),
    axis.ticks = ggplot2::element_line(),
    axis.text.x = ggplot2::element_text(size = 10),
    axis.text.y = ggplot2::element_text(size = 10),
    axis.title = ggplot2::element_text(size = 11),
    plot.title = ggplot2::element_text(size = 12, face = "bold"),
    axis.ticks.length = grid::unit(0.25, "cm"),
    legend.title = ggplot2::element_text(size = 11),
    legend.text = ggplot2::element_text(size = 10)
  )

  vcv_plots <- lapply(names(vcv_list), function(name) {
    ggplot2::ggplot(melt_matrix(vcv_list[[name]]), ggplot2::aes(Trait1, Trait2, fill = Value)) +
      ggplot2::geom_tile() +
      viridis::scale_fill_viridis(
        name = "VCV Value",
        option = "D",
        limits = vcv_limits
      ) +
      ggplot2::labs(title = name, x = NULL, y = NULL) +
      ggplot2::theme_minimal(base_size = 11) +
      clean_heatmap_theme
  })

  cor_plots <- lapply(names(cor_list), function(name) {
    ggplot2::ggplot(melt_matrix(cor_list[[name]]), ggplot2::aes(Trait1, Trait2, fill = Value)) +
      ggplot2::geom_tile() +
      viridis::scale_fill_viridis(
        name = "Correlation",
        option = "B",
        limits = c(0, 0.75)
      ) +
      ggplot2::labs(title = name, x = NULL, y = NULL) +
      ggplot2::theme_minimal(base_size = 11) +
      clean_heatmap_theme
  })

  eigen_df <- do.call(rbind, list(
    data.frame(
      Eigenvalue = eigen(ancestral_sigma)$values,
      Rank = seq_len(n_traits),
      Scenario = "Ancestral"
    ),
    data.frame(
      Eigenvalue = eigen(integration_rescaled$rescaled_sigma)$values,
      Rank = seq_len(n_traits),
      Scenario = "Integration-rate"
    ),
    data.frame(
      Eigenvalue = eigen(proportional_sigma)$values,
      Rank = seq_len(n_traits),
      Scenario = "Proportional-scaled"
    )
  ))

  eff_dim_df <- data.frame(
    Scenario = c("Ancestral", "Integration-rate", "Proportional-scaled"),
    EffDimensionality = c(
      compute_effective_dimensionality(ancestral_sigma),
      compute_effective_dimensionality(integration_rescaled$rescaled_sigma),
      compute_effective_dimensionality(proportional_sigma)
    )
  )

  mean_var_df <- data.frame(
    Scenario = c("Ancestral", "Integration-rate", "Proportional-scaled"),
    MeanVariance = c(
      mean(diag(ancestral_sigma)),
      mean(diag(integration_rescaled$rescaled_sigma)),
      mean(diag(proportional_sigma))
    )
  )

  p_effdim <- ggplot2::ggplot(eff_dim_df, ggplot2::aes(x = Scenario, y = EffDimensionality, fill = Scenario)) +
    ggplot2::geom_bar(stat = "identity") +
    ggplot2::labs(title = "Effective Dimensionality", y = "Effective Dimensionality", x = NULL) +
    ggplot2::theme_minimal(base_size = 11) +
    ggplot2::scale_fill_viridis_d() +
    l_shaped_theme +
    ggplot2::theme(
      axis.title.x = ggplot2::element_blank(),
      axis.text.x = ggplot2::element_blank(),
      legend.position = "none"
    )

  p_eig <- ggplot2::ggplot(eigen_df, ggplot2::aes(x = Rank, y = Eigenvalue, color = Scenario)) +
    ggplot2::geom_line(linewidth = 1.2) +
    ggplot2::labs(title = "Eigenvalue Spectra", x = "Rank", y = "Eigenvalue") +
    ggplot2::theme_minimal(base_size = 11) +
    ggplot2::scale_color_viridis_d() +
    l_shaped_theme +
    ggplot2::theme(legend.position = "none")

  p_trace <- ggplot2::ggplot(mean_var_df, ggplot2::aes(x = Scenario, y = MeanVariance, fill = Scenario)) +
    ggplot2::geom_bar(stat = "identity") +
    ggplot2::labs(title = "Mean Diagonal Variance", y = "Mean Variance", x = NULL) +
    ggplot2::theme_minimal(base_size = 11) +
    ggplot2::scale_fill_viridis_d() +
    l_shaped_theme +
    ggplot2::theme(
      axis.title.x = ggplot2::element_blank(),
      axis.text.x = ggplot2::element_blank()
    )

  vcv_plots[[1]] <- vcv_plots[[1]] + ggplot2::theme(legend.position = "none")
  vcv_plots[[2]] <- vcv_plots[[2]] + ggplot2::theme(legend.position = "none")
  cor_plots[[1]] <- cor_plots[[1]] + ggplot2::theme(legend.position = "none")
  cor_plots[[2]] <- cor_plots[[2]] + ggplot2::theme(legend.position = "none")

  title_text <- "Empirically Calibrated Simulation Scenarios"

  ((vcv_plots[[1]] + vcv_plots[[2]] + vcv_plots[[3]]) /
     (cor_plots[[1]] + cor_plots[[2]] + cor_plots[[3]]) /
     (p_effdim + p_eig + p_trace)) +
    patchwork::plot_annotation(
      title = title_text,
      tag_levels = "A",
      theme = ggplot2::theme(
        plot.title = ggplot2::element_text(size = 14, face = "bold", hjust = 0.5),
        plot.tag = ggplot2::element_text(size = 13, face = "bold")
      )
    )
}

simulation_scenarios_caption_text <- paste0(
  "Empirically calibrated simulation scenarios. Panels A-C show the ancestral VCV, ",
  "proportional scaling with s = 0.5, and a reduced-integration/faster-rate ",
  "shift with g = 0.65. Panels D-F show their correlation matrices. Panels ",
  "G-I summarize correlation effective dimensionality, covariance eigenvalue ",
  "spectra, and mean marginal evolutionary variance. The proportional scenario ",
  "changes scale without changing correlation structure; the integration-rate ",
  "scenario changes covariance shape and rate together."
)
```

```{r, echo = FALSE, results = "asis"}
if (knitr::is_html_output()) cat("
<style>
img, .figure img, .figure > p > img {
  border: 0 !important;
  box-shadow: none !important;
  outline: none !important;
  background: transparent !important;
}

p.caption, .caption, table > caption {
  border-left: 0.32rem solid #d8dde3;
  color: #2f3439;
  font-size: inherit;
  line-height: inherit;
  margin: 0 3.8% 1.05rem 3.8%;
  padding: 0 0 0 1rem;
  text-align: left !important;
}

p.caption strong, .caption strong {
  font-weight: 700;
}

table.simulation-preview-table > caption {
  caption-side: bottom !important;
  margin: 0.65rem 3.8% 1.35rem 3.8%;
}

table.simulation-preview-table > caption strong {
  font-weight: 700;
}

table {
  width: 100%;
}

table th, table td {
  vertical-align: top;
}

table code {
  white-space: nowrap;
}

.simulation-table-wrap {
  margin: 1em 0 1.15em 0;
  max-width: 100%;
  overflow-x: auto;
}

.simulation-table-wrap table {
  margin-bottom: 0;
}

.simulation-preview-table thead th {
  white-space: nowrap;
}

@media (max-width: 600px) {
  h1.title {
    font-size: 2rem;
    line-height: 1.1;
  }
}
</style>
")
```

> **Development note:** This simulation and tuning workflow is under active development and remains somewhat experimental. The interface and recommendations may change as it is evaluated more broadly.

This is Part 1 of the empirically calibrated simulation workflow. Here we examine performance at fixed search settings; [Part 2](simulation-study-part-2.html) compares alternative settings on the same simulated datasets and shows how to pass selected settings to an empirical search.

How well does `bifrost` recover shifts in datasets resembling your own? Here we assess performance using simulations calibrated to the empirical tree, trait dimensionality, fitted means, and residual covariance scale.

We then compare search behavior under three scenarios:

1. a null model with no shifts, to estimate expected false positives,
2. proportional covariance shifts, which match the fitted model, and
3. an integration-rate scenario that departs from the fitted model, coupling reduced integration with faster marginal rates, or increased integration with slower marginal rates.

These simulations can guide search settings before an analysis or assess their performance afterward. Repeated subtree resampling lets us explore different topologies and parameter draws without searching the full tree in every replicate.

We use the passerine body-plan data from Berv et al. (2026) to calibrate this worked example. The simulations shown here are a new tutorial exercise, not the analyses reported in that study.

The saved summaries reproduce the tables below without fitting models or running simulations; the simulation examples are optional.

Start by loading `bifrost`.

```{r load_packages, eval = FALSE}
# Load bifrost before constructing templates or running simulation studies.
library(bifrost)
```

This part requests the passerine tree (`passerine-tree`, approximately 81 KB),
trait matrix (`passerine-traits`, approximately 212 KB), and simulation preview
tables (`simulation-preview-tables`, approximately 6 KB). First use downloads
and verifies each file; later calls reuse the verified cache. The larger tree
and trait files share that resolver cache with the passerine articles. This
explicit command checks for an updated preview table:

```{r refresh_simulation_preview_tables, eval=FALSE}
# Refresh the preview tables only when an updated copy is wanted.
bifrost_example_file("simulation-preview-tables", refresh = TRUE)
```

## Build an empirically calibrated template

Here we use the empirical calibration structure from Berv et al. (2026), which models 12 skeletal traits jointly while keeping log body mass as a phylogenetic covariate.

```{r load_bodyplan_example}
# Resolve the passerine tree and calibrated skeletal measurements.
tree_path <- bifrost_example_file("passerine-tree")
trait_path <- bifrost_example_file("passerine-traits")

# Align the trait rows to the tree's tip order.
bird_tree <- ape::read.tree(tree_path)
bodyplan_data <- readRDS(trait_path)
bodyplan_data <- as.matrix(bodyplan_data[bird_tree$tip.label, , drop = FALSE])

# Keep body mass as the predictor and the skeletal variables as responses.
skeletal_cols <- setdiff(colnames(bodyplan_data), "vertnet_mass")
bodyplan_data <- bodyplan_data[, c(skeletal_cols, "vertnet_mass"), drop = FALSE]

# Express the multivariate response and body-mass covariate by column index.
formula_str <- sprintf(
  "trait_data[, 1:%d] ~ trait_data[, %d]",
  length(skeletal_cols),
  ncol(bodyplan_data)
)

# Confirm the aligned sample size and response dimensionality.
c(
  species = nrow(bodyplan_data),
  skeletal_traits = length(skeletal_cols),
  total_columns = ncol(bodyplan_data)
)
```

We fit a single-regime multivariate BM model with `mvMORPH::mvgls()` using `method = "LL"`, following [Clavel, Aristide, and Morlon (2019)](https://doi.org/10.1093/sysbio/syy045) and [Clavel and Morlon (2020)](https://doi.org/10.1093/sysbio/syaa010). The template retains the aligned data, fitted model, means, and covariance summaries used to generate a fresh ancestral covariance matrix for each replicate.

```{r create_template_code, eval = FALSE}
# Opt in only when fitting a template and running new simulation studies.
run_simulations <- FALSE
if (run_simulations) {
  set.seed(5, kind = "Mersenne-Twister")
  bodyplan_template <- createSimulationTemplate(
    baseline_tree = bird_tree,
    trait_data = bodyplan_data,
    formula = formula_str,
    response_columns = seq_along(skeletal_cols),
    predictor_columns = ncol(bodyplan_data),
    method = "LL",
    error = TRUE
  )
  bodyplan_template
}
```

You can skip this fit when reading the saved results. To run the individual studies below interactively or in Colab, change `run_simulations` to `TRUE` in the block above before executing it. The empirical calibration uses `error = TRUE`; searches on simulated responses use `error = FALSE`.

Covariates such as body mass shape the calibration fit, but the studies simulate only the response traits and use intercept-only shift searches. Without covariates, use the simpler `trait_data ~ 1` template. Response columns must be raw numeric variables that can be regenerated in each replicate.

The examples below use `simulation_generator = "empirical"`, the recommended option when the fitted full covariance structure should anchor new draws.

## Simulation scenarios

The figure below places the three package scenarios side by side: an empirical ancestral covariance matrix, a proportional rescaling, and the integration-rate trade-off. The last scenario raises the eigenvalues of the ancestral correlation matrix to a power `g`, renormalizes to a correlation matrix, and multiplies the covariance matrix by `1 / g`. Thus `g < 1` reduces integration while increasing every marginal evolutionary variance; `g > 1` increases integration while decreasing every marginal variance.

```{r simulation_scenarios_figure, echo = FALSE, fig.width = 8.8, fig.height = 8.4, out.width = "100%", fig.alt = "Schematic of the empirically calibrated null, proportional, and integration-rate simulation scenarios.", fig.cap = vignette_figure_caption(paste0("**Figure 1.** ", simulation_scenarios_caption_text), simulation_scenarios_caption_text)}
build_simulation_scenario_figure(
  seed = 123,
  n_traits = 10,
  var_mean = 1,
  var_sd = 0.1,
  cov_mean = 0.5,
  cov_sd = 0.25,
  proportional_scale = 0.5,
  integration_power = 0.65
)
```

In the proportional panels, the covariance structure keeps the same shape and simply expands or contracts. In the integration-rate panels, weaker correlation structure is paired with larger marginal variance, so covariance shape and evolutionary rate change together. The shifted passerine examples below use these two simulation styles under repeated subtree resampling from the full tree, alongside a null no-shift baseline.

```{r preview_results_setup, echo = FALSE, message = FALSE, warning = FALSE}
preview_tables_path <- bifrost_example_file("simulation-preview-tables")
preview_tables <- readRDS(preview_tables_path)
if (!identical(preview_tables$schema_version, 4L)) {
  stop(paste0(
    "The cached simulation preview tables are stale; run ",
    'bifrost_example_file("simulation-preview-tables", refresh = TRUE).'
  ))
}
stopifnot(
  identical(
    preview_tables$provenance$fixed_settings$simulation_generator,
    "empirical"
  ),
  identical(
    preview_tables$provenance$fixed_settings$integration_power_range,
    c(0.5, 1.25)
  ),
  identical(
    preview_tables$provenance$fixed_settings$integration_exclude_range,
    c(0.8, 1.1)
  ),
  identical(preview_tables$provenance$fixed_settings$n_replicates_per_setting, 500L),
  identical(preview_tables$provenance$fixed_settings$min_shift_tips, 10L),
  identical(preview_tables$provenance$fixed_settings$max_shift_tips, 40L),
  identical(preview_tables$provenance$fixed_settings$search_error, FALSE)
)
```

## Before running simulations

Keep three practical details in mind.

- Use the study function's `seed` argument for reproducibility. Do not pass a separate seed inside `simulation_options`; the study functions manage replicate-level seeds.

- Study functions inherit template fit settings such as `method` and `error`, so we explicitly override `error` for simulated searches. Parallelize either the study or each search, not both; these examples run replicates in parallel and keep each search serial.

- Distinguish a successful search that finds no shifts from one with no eligible candidates or a failed fit. A successful zero-shift search is included in recovery evaluation: its true shifts count as missed. Check `per_replicate$n_candidates`; with no candidates, recall can still be evaluated, but candidate-based specificity and false-positive rate are undefined for that replicate. Fitting failures are reported separately. The cutoff changes which nodes can be tested, not just the support required to accept a shift.

The next three sections show the individual study calls. The fixed-settings comparison then uses cached preview results to show the same logic side by side.

```{r simulation_figure_page, echo = FALSE, results = "asis"}
# Flush the full-page figure before any simulation code begins.
if (knitr::is_latex_output()) cat("\\clearpage\n")
```

## 1. Estimate False Positives

The null study asks the baseline error question: if the real dataset contained no shifts at all, how often would the search still infer one?

```{r false_positive_study, eval = FALSE}
# Estimate false positives under an empirical-covariance null model.
fp_study <- if (run_simulations) runFalsePositiveSimulationStudy(
  template = bodyplan_template,
  n_replicates = 500,
  tree_tip_count = 250,
  simulation_options = list(
    simulation_generator = "empirical"
  ),
  search_options = list(
    formula = "trait_data ~ 1",
    min_descendant_tips = 10,
    shift_acceptance_threshold = 10,
    IC = "GIC",
    num_cores = 1,
    method = "LL",
    error = FALSE,
    ic_uncertainty_threshold = 10,
    plot = FALSE,
    store_model_fit_history = FALSE
  ),
  num_cores = 8,
  seed = 5
)
```

```{r false_positive_inspect, eval = FALSE}
# Inspect the study object, representative replicates, and aggregate summary.
if (run_simulations) {
  fp_study
  head(fp_study$per_replicate)
  fp_study$study_summary
}
```

Null simulations are useful for screening candidate choices of `shift_acceptance_threshold`, `IC`, and `min_descendant_tips`.

## 2. Estimate Recovery Under Proportional Scaling

Recovery under proportional scaling evaluates the scenario that matches the current `bifrost` BMM search assumptions: shifted covariance matrices are scalar multiples of the background covariance matrix.

```{r proportional_recovery_study, eval = FALSE}
# Measure recovery when simulated shifts match proportional BMM scaling.
recovery_prop <- if (run_simulations) runShiftRecoverySimulationStudy(
  template = bodyplan_template,
  n_replicates = 500,
  tree_tip_count = 250,
  simulation_options = list(
    simulation_generator = "empirical",
    num_shifts = 5,
    min_shift_tips = 10,
    max_shift_tips = 40,
    scale_mode = "proportional",
    scale_factor_range = c(0.1, 2.0),
    exclude_range = c(0.5, 1.5),
    buffer = 3
  ),
  search_options = list(
    formula = "trait_data ~ 1",
    min_descendant_tips = 10,
    shift_acceptance_threshold = 10,
    IC = "GIC",
    num_cores = 1,
    method = "LL",
    error = FALSE,
    ic_uncertainty_threshold = 10,
    plot = FALSE,
    store_model_fit_history = FALSE
  ),
  fuzzy_distance = 2,
  weighted = TRUE,
  num_cores = 8,
  seed = 5
)
```

```{r proportional_recovery_inspect, eval = FALSE}
# Compare strict, fuzzy, and distance-weighted recovery summaries.
if (run_simulations) {
  recovery_prop
  recovery_prop$study_summary
  recovery_prop$evaluation$strict
  recovery_prop$evaluation$fuzzy
  recovery_prop$evaluation$weighted
}
```

## 3. Check the Integration-Rate Robustness Scenario

It is also useful to test a deliberately mismatched scenario in which trait integration and evolutionary rate evolve together. This asks how well the same search settings behave when the data depart from strict proportional VCV scaling. The moderate power bands below sample reduced integration from `0.5-0.8` and increased integration from `1.1-1.25`; each power `g` scales marginal variances by exactly `1 / g`.

```{r integration_recovery_study, eval = FALSE}
# Stress-test the same search under integration-rate model misspecification.
recovery_integration <- if (run_simulations) runShiftRecoverySimulationStudy(
  template = bodyplan_template,
  n_replicates = 500,
  tree_tip_count = 250,
  simulation_options = list(
    simulation_generator = "empirical",
    num_shifts = 5,
    min_shift_tips = 10,
    max_shift_tips = 40,
    scale_mode = "correlation",
    integration_power_range = c(0.5, 1.25),
    integration_exclude_range = c(0.8, 1.1),
    buffer = 3
  ),
  search_options = list(
    formula = "trait_data ~ 1",
    min_descendant_tips = 10,
    shift_acceptance_threshold = 10,
    IC = "GIC",
    num_cores = 1,
    method = "LL",
    error = FALSE,
    ic_uncertainty_threshold = 10,
    plot = FALSE,
    store_model_fit_history = FALSE
  ),
  fuzzy_distance = 2,
  weighted = TRUE,
  num_cores = 8,
  seed = 5
)
```

```{r integration_recovery_inspect, eval = FALSE}
# Review the overall study and the fuzzy recovery result used for reporting.
if (run_simulations) {
  recovery_integration
  recovery_integration$study_summary
  recovery_integration$evaluation$fuzzy
}
```

This scenario changes covariance shape while the fitted search model permits only proportional covariance shifts. It therefore explores robustness to model misspecification. Recovery also depends on the magnitudes of the simulated rate changes, which differ between these scenarios; the comparison does not isolate the effect of covariance shape alone.

## 4. Compare Fixed Settings Across Scenarios

We now compare GIC and BIC at the same fixed settings across the null, proportional, and integration-rate scenarios. The saved results use 500 datasets per scenario, each based on a freshly resampled 250-tip subtree of the passerine tree with 12 simulated response traits.

The comparison holds `shift_acceptance_threshold = 10` and `min_descendant_tips = 10`. Each shifted dataset contains five planted shifts in clades with 10 to 40 descendant tips. These settings illustrate performance before tuning; Part 2 compares them with more restrictive settings on the same datasets.

The individual calls above demonstrate each study separately. The saved tables instead extract the 10/10 settings from the paired grid in [Part 2](simulation-study-part-2.html), where every setting is evaluated on the same datasets. Use that paired workflow and its reproduction scripts to reproduce these results; running the individual studies with their own seeds will generate different draws.

```{r fixed_settings_cached}
# Read the compact summaries without rerunning searches.
preview_tables <- readRDS(bifrost_example_file("simulation-preview-tables"))
fixed_settings_table <- preview_tables$fixed_settings
```

```{r medium_case_study_cached, echo = FALSE, message = FALSE, warning = FALSE}
fixed_null_display <- fixed_settings_table[
  fixed_settings_table$Scenario == "Null",
  c("IC", "Mean FP", "Any FP", "Mean shifts"),
  drop = FALSE
]
fixed_null_display[c("Mean FP", "Any FP")] <-
  100 * fixed_null_display[c("Mean FP", "Any FP")]
names(fixed_null_display)[2:3] <- c("Null FP (%)", "Null any FP (%)")
fixed_recovery_display <- fixed_settings_table[
  fixed_settings_table$Scenario != "Null",
  c(
    "IC", "Scenario", "Fuzzy recall", "Fuzzy specificity", "Fuzzy F1",
    "Fuzzy balanced accuracy", "Mean shifts"
  ),
  drop = FALSE
]

fixed_metric <- function(ic, scenario, metric) {
  value <- fixed_settings_table[
    fixed_settings_table$IC == ic & fixed_settings_table$Scenario == scenario,
    metric,
    drop = TRUE
  ]
  stopifnot(length(value) == 1L, is.finite(value))
  formatC(value, digits = 3, format = "f")
}
```

Table 1 summarizes false positives under the null; Table 2 summarizes recovery in datasets with planted shifts.

`Null FP (%)` is the mean percentage of eligible candidate nodes incorrectly assigned a shift. `Null any FP (%)` is the percentage of null datasets with at least one inferred shift. All 500 searches per scenario and IC completed successfully, including searches that inferred no shifts.

```{r medium_case_null_table, echo = FALSE, results = "asis"}
render_preview_table(
  fixed_null_display,
  caption = "Table 1. Null behavior at an acceptance threshold of 10 and minimum clade size of 10 across 500 repeated 250-tip subsamples.",
  col_names = c("IC", "Null FP (%)", "Null any FP (%)", "Mean shifts"),
  digits = c(0, 3, 1, 2)
)
```

```{r medium_case_recovery_table, echo = FALSE, results = "asis"}
render_preview_table(
  fixed_recovery_display,
  caption = "Table 2. Shift recovery for the same fixed settings and repeated subsamples. Fuzzy scoring credits near misses within two nodes of a true shift. Balanced accuracy is the mean of fuzzy recall and fuzzy specificity; F1 is retained as the complementary precision-recall summary.",
  col_names = c(
    "IC", "Scenario", "Fuzzy rec.", "Fuzzy spec.", "Fuzzy F1",
    "Fuzzy BA", "Mean shifts"
  ),
  digits = c(0, 0, 3, 3, 3, 3, 2)
)
```

Because unshifted candidate nodes usually outnumber true shifts, raw accuracy can look high even when recovery is poor. Balanced accuracy gives recall and specificity equal weight; F1 instead combines precision and recall. Fuzzy scoring credits nearby inferred shifts as matches, allowing for uncertainty in their exact location.

For GIC, fuzzy balanced accuracy is `r fixed_metric("GIC", "Proportional", "Fuzzy balanced accuracy")` under proportional shifts and `r fixed_metric("GIC", "Integration-rate", "Fuzzy balanced accuracy")` under the integration-rate trade-off. The corresponding BIC values are `r fixed_metric("BIC", "Proportional", "Fuzzy balanced accuracy")` and `r fixed_metric("BIC", "Integration-rate", "Fuzzy balanced accuracy")`. GIC recovers more shifts at these settings, but also infers shifts more often under the null. Part 2 shows how that trade-off affects the choice of search settings.

## Performance Summary

These checks illustrate false-positive behavior and shift recovery under the simulated conditions, rather than establishing settings that will work for every dataset. [Part 2](simulation-study-part-2.html) expands this fixed-setting comparison into a paired tuning grid on the same datasets and shows how to use the selected settings in an empirical search.

## References

If you use these data or reproduce this workflow, the most relevant citations are:

- Berv, Jacob S., Charlotte M. Probst, Santiago Claramunt, J. Ryan Shipley, Matt Friedman, Stephen A. Smith, David F. Fouhey, and Brian C. Weeks. 2026. "Rates of passerine body plan evolution in time and space." *Nature Ecology & Evolution*. <https://doi.org/10.1038/s41559-026-03110-5>
- Berv, Jacob S., Charlotte M. Probst, Santiago Claramunt, J. Ryan Shipley, Matt Friedman, Stephen A. Smith, David F. Fouhey, and Brian C. Weeks. 2026. "Supplementary data archive for Rates of passerine body plan evolution in time and space" (v1.0.0) [Data set]. Zenodo. <https://doi.org/10.5281/zenodo.19198393>
- Clavel, Julien, Laurent Aristide, and Helene Morlon. 2019. "A penalized likelihood framework for high-dimensional phylogenetic comparative methods and an application to New-World monkeys brain evolution." *Systematic Biology* 68(1):93-116. <https://doi.org/10.1093/sysbio/syy045>
- Clavel, Julien, and Helene Morlon. 2020. "Reliable phylogenetic regressions for multivariate comparative data: Illustration with the MANOVA and application to the effect of diet on mandible morphology in phyllostomid bats." *Systematic Biology* 69(5):927-943. <https://doi.org/10.1093/sysbio/syaa010>
- Clavel, Julien, Gilles Escarguel, and Gildas Merceron. 2015. "mvMORPH: an R package for fitting multivariate evolutionary models to morphometric data." *Methods in Ecology and Evolution* 6(11):1311-1319. <https://doi.org/10.1111/2041-210X.12420>
- Smith, Stephen A., Nicholas Walker-Hale, and Caroline T. Parins-Fukuchi. 2023. "Compositional shifts associated with major evolutionary transitions in plants." *New Phytologist* 239(6):2404-2415. <https://doi.org/10.1111/nph.19099>

## Software Used in This Vignette

- `bifrost` for empirical simulation templates, null and shifted datasets, repeated simulation studies, recovery evaluation, and tuning-grid summaries.
- `mvMORPH`/`mvgls` machinery underlies the global calibration fit and empirical shift searches.
- `ape` and `phytools` underlie the phylogenetic tree, SIMMAP, and subtree operations used by the simulation helpers.
- `ggplot2`, `patchwork`, and `viridis` draw the schematic simulation-scenario figure.
- `knitr` and `rmarkdown` render the vignette and preview tables.

## AI Assistance

This vignette was developed with assistance from OpenAI tools for drafting, editing, and figure refinement; all scientific content, interpretation, and final decisions were reviewed by the authors.
